The causes and consequences of human-specific DNA methylation

Version 1

Abstract

The vast collection of human-specific traits– such as our unique morphology, cognition, behavior, and diseases– has long been attributed to gene expression divergence between us and our closest living relatives, chimpanzees. Theory suggests that changes to cis-regulatory elements such as promoters and enhancers may drive evolutionary adaptation, and DNA methylation is a key factor in transcriptional cis-regulation. However, we still lack an understanding of 1) how species-specific methylation patterns arise; 2) their downstream effects; and 3) whether they are a common target of natural selection. In this study, we investigated these three questions. By combining a novel hypothesis testing framework with DNA methylation data from six human and chimpanzee cell types, as well as fused interspecies hybrid cells, we disentangled cis- vs. trans-acting methylation divergence across the genome. Across cell types, we found that methylation divergence is primarily driven in cis, which can be linked in some cases to nearby sequence variants such as CpG gains and losses. Although less common, regions with trans-acting methylation divergence were enriched for specific transcription factor (TF) binding motifs, suggesting a role of TFs such as FOXM1 in these differences. Having established these causes of methylation divergence, we then examined the functional consequences of differential methylation. Although methylation lacks a consistent relationship with transcription, we observed that associations between methylation and gene expression are stronger for genes with cis-regulatory divergence. Moreover, we identified lineage-specific selection shaping promoter methylation at the level of entire pathways including those affecting human-specific traits such as speech, cognition, and susceptibility to infection with hepatitis C. Collectively, our findings provide a mechanistic framework suggesting that DNA methylation may occupy a key position, mediating the effects of both cis- and trans-acting factors on transcriptional networks, including those contributing to human-specific traits.

Introduction

While gene expression divergence has long been considered the primary driver of human evolution, identifying the molecular mechanisms underlying uniquely human traits remains challenging, particularly when they involve epigenetic modifications that control cell type-specific expression patterns1–4. Theory suggests that cell type-specific cis-regulatory variants may fuel evolutionary adaptation through their precise effects on individual genes within specific cell types, avoiding potentially deleterious consequences of broader trans-acting changes that affect multiple targets simultaneously5,6. However, while recent advances have enabled quantification of human-specific divergence in gene expression and chromatin accessibility, the role of DNA methylation—a key regulator of spatiotemporal gene expression—in coordinating these changes and driving human-specific traits has not been fully explored6–10. DNA methylation, occurring primarily at CpG dinucleotides, is generally associated with transcriptional repression and provides stable, long-term regulatory control that has been linked to species-specific phenotypes in mammals11. Comparative methylome studies between humans and chimpanzees have revealed widespread tissue-specific methylation divergence, and linked human-specific hypomethylation to neurodevelopmental evolution and disease susceptibility7,12–23.

Studies of allele-specific expression (ASE) and chromatin accessibility measurements in interspecies hybrids provide a powerful approach for studying regulatory evolution by eliminating confounding factors inherent to cross-species comparisons, including differences in cell type composition, environmental conditions, and developmental timing24–33. Analogously, allele-specific methylation (ASM) in human-chimpanzee hybrid cells enables the isolation of cis-regulatory effects while controlling for shared trans-acting factors. Cis regulation of methylation refers to methylation differences caused by factors such as sequence variants on the same allele, typically acting locally. In contrast, trans regulation of methylation involves diffusible factors such as DNA methyltransferases, transcription factors, or chromatin modifiers that can influence methylation patterns across many genomic loci (Figure 2A). Studying ASM enables direct quantification of the relative contribution of local versus diffusible factors to species-specific methylation patterns.

While previous studies of primate methylation evolution in post-mortem tissues and blood have provided valuable foundational insights7,12–16,18–23,34,35, several important questions remain to be addressed. These include: (1) generating comprehensive genome-wide methylation maps across multiple species and cell types; (2) distinguishing between cis- versus trans-regulatory mechanisms underlying species-specific methylation differences; (3) identifying cis-acting genetic variants and trans-acting factors underlying these differences; and (4) linking natural selection on methylation to gene expression and candidate traits. Addressing these gaps would enhance our understanding of how genetic variants impact methylation and in turn shape adaptive traits.

Human-chimpanzee tetraploid hybrid induced pluripotent stem cells (iPSCs), generated through in vitro cell fusion, provide a controlled experimental system for quantifying cis- and trans-regulatory contributions to methylation divergence by hosting both species’ alleles in an identical trans-environment. By measuring ASM and ASE across multiple cell types, this system generates genome-wide maps of cell type-specific regulatory differences that may contribute to evolutionary adaptation. In this study, we generated bisulfite sequencing data from human, chimpanzee, and hybrid iPSCs, cranial neural crest cells, dopaminergic neurons, skeletal myocytes, and hepatocytes, as well as from primary human and chimpanzee dental pulp stem cells. We combined these data with previously published27 and newly generated RNA-seq data from all six cell types.

Using these data, we investigated the causes of methylation divergence and quantified contributions of cis- and trans-acting mechanisms that establish interspecies differences in methylation patterns. We developed a hypothesis testing framework tailored to count-based proportional methylation data and extended a changepoint detection algorithm to identify differentially methylated regions (DMRs) under consistent cis- or trans-regulation. We show that cis-acting mechanisms predominantly drive genome-wide methylation divergence, with cis-regulated DMRs likely affected by nearby species-specific single-nucleotide variants. In contrast, trans-regulated DMRs were enriched for transcription factor binding motifs, including pioneer factors and chromatin remodelers that may mediate trans-acting divergence. Concordance between ASM and ASE across cell types revealed species-specific and context-dependent regulatory relationships linking methylation to gene expression. Finally, coordinated methylation-expression divergence in gene sets related to speech, cognition, and viral infection susceptibility connected lineage-specific selection to candidate phenotypes. In sum, our work provides a framework for understanding how divergent cis- and trans-acting factors converge on methylation patterns to produce human-specific traits.

Results

Measurement of DNA methylation across diverse cell types

To establish comprehensive methylation landscapes across evolutionarily relevant cell types, we differentiated human, chimpanzee, and their fused tetraploid hybrid iPSCs into four cell types (see Methods)25,27. We included an additional cell type, dental pulp stem cells, which are primary cells isolated from human and chimpanzee donors25. We then generated whole-genome bisulfite sequencing (WGBS) data from human and chimpanzee parental and hybrid cranial neural crest cells (CNCC) and dopaminergic neurons, as well as parental dental pulp stem cells (DPSC), complemented by reduced representation bisulfite sequencing (RRBS) data from induced pluripotent stem cells (iPSCs), skeletal myocytes (SKM) and hepatocytes (HEP) (Figure 1A). Principal component analysis and hierarchical clustering of methylation patterns showed that samples clustered primarily by cell type, then by species, and finally by parental versus hybrid system (Figure 1B and 1C and Figure 1—figure supplement 1). CNCC samples displayed a unique clustering pattern where parental vs. hybrid samples clustered separately, though examination of the dendrogram indicated this was due to a single human parental sample being a slight outlier (Figure 1B). Known cell type-specific gene expression markers validated the correct identity of each differentiated cell type (Figure 1D and Figure 1—figure supplement 2)36–50. Overall, these results are consistent with our expectation that cell type is the predominant determinant of DNA methylation patterns, while species explains a much smaller proportion7. In other words, a human hepatocyte is much more similar to a chimpanzee hepatocyte than it is to a human dopaminergic neuron.

DNA methylation profiles across diverse human, chimpanzee, and hybrid cell types.

(A) Four cell types were differentiated from human, chimpanzee, and hybrid induced pluripotent stem cells. The cell types span diverse body systems including skeletal myocytes for the musculoskeletal system, dopaminergic neurons for the central nervous system, cranial neural crest cells for craniofacial development and hepatocyte progenitors for the liver. We collected DNA methylation and RNA-seq data from these cell types from humans, chimpanzees and their hybrids. Dental pulp stem cells were also collected from human and chimpanzee adult tissues. (B) Clustering analysis and (C) PCA between samples for all cell types (left) and for all iPSC-derived cell types (i.e. excluding DPSC, right). (D) Cell type specific markers of gene expression. Figure 1A was created using BioRender, and is published under a CC BY-NC-ND 4.0 license.

Contribution of cis and trans regulation to DNA methylation divergence between human and chimpanzee

To understand the molecular mechanisms underlying species-specific methylation patterns, we first established a framework for distinguishing between cis- and trans-regulatory control of DNA methylation. By comparing methylation levels between human and chimpanzee alleles within the same cellular environment (hybrid cells), we can isolate cis effects from trans effects and determine which mechanism predominates in establishing species-specific methylation patterns (Figure 2A).

Contribution of cis and trans regulation to DNA methylation divergence between human and chimpanzee.

(A) Interspecies differences in methylation levels per CpG site is compared between parental and hybrid systems and assigned a regulation category based on hypothesis testing. (B) Parental and hybrid methylation ratios in dopaminergic neurons and the respective regulation category (color key in panel C). (C) Quantification of relative contribution of non-conserved regulation groups to whole-genome CpG sites across cell types. (D) Heterogeneity of the distribution of CpG sites of all regulation groups. (E) Quantification of CpG regulatory clusters by overlap with genomic features of transcripts genome-wide as well as cis-regulatory elements including promoters, enhancers and CTCF-bound cis-regulatory regions (CTCF). (F) Quantification of relative contribution of non-conserved regulation groups to regulatory clusters across cell types (color key in panel C). Figure 2A was created using BioRender, and is published under a CC BY-NC-ND 4.0 license.

To quantify the relative contributions of cis- versus trans-acting factors in establishing human-chimpanzee methylation differences, we applied a beta-binomial model framework adapted from Hallgrímsdóttir et al. (2024) to classify differentially methylated regions based on allele-specific patterns in hybrid versus parental samples. To determine whether a CpG site reflects a difference in methylation due to cis or trans regulation, two hypothesis tests are used: one to evaluate the null hypothesis that the difference in methylation is due solely to trans, and the other solely to cis (see Methods). Each site is categorized into one of five mutually exclusive regulation groups– pure cis-driven, pure trans-driven, cis + trans, and cis x trans, and conserved (Figure 2A). We investigated genome-wide methylation divergence by fitting the model to methylation levels at each CpG site. In all cell types examined, most CpG sites (83-93%) had conserved patterns of methylation. Among divergent sites, cis-regulation was over twice as common as trans-regulation, while a smaller fraction showed both cis and trans divergence (Figure 2B, 2C, Figure 2—figure supplement 1 and Supplemental File 1). The predominance of cis- over trans-regulation was maintained after FDR correction and remained stable across a range of FDR cutoffs (FDR < 0.25, 0.20, 0.10, 0.05, 0.01; Figure 2—figure supplement 1-2 and Supplemental Files 1-2). This genome-wide predominance of cis-regulation indicates that local sequence differences between human and chimpanzee genomes are the primary drivers of methylation divergence, consistent with previous hypotheses of gene regulation that cis-regulatory changes may be less disruptive than broad trans-acting modifications1–4,6.

The regional distribution of sites from different regulation groups reveals a consistent pattern across cell types: conserved CpG sites are most likely to be near other conserved sites, while divergent CpG sites are likely to be near conserved sites as well as sites within the same regulation group (Figure 2D). This suggests that CpG sites within the same regulation group tend to form coherent clusters in the genome, and divergent CpG clusters typically have conserved sites scattered within. When conducting regional analyses, methylation data is usually pooled from multiple sites within a pre-defined genomic region6–10. However, this regional heterogeneity can affect classification as well as estimates of how much each regulatory mechanism contributes to divergence between species (Figure 2—figure supplement 2-3, Supplemental File 2, see Methods). Therefore, pooling data across regions should be approached carefully, as subsets of CpG sites within the same regulatory element can be controlled by distinct mechanisms. These findings underscore the importance of studying methylation divergence at high resolution.

To identify differentially methylated regions (DMRs) exhibiting consistent inter-species methylation directionality and minimal classification heterogeneity of regulation groups, we implemented a changepoint detection method with a tolerance threshold for regulatory homogeneity (see Methods). This method systematically identifies genomic loci where co-localized CpG sites demonstrate uniform regulatory behavior punctuated by a pre-defined fraction of heterogeneous sites: non-conserved classifications showing consistent directional methylation bias, and conserved sites maintaining equal methylation levels between species. This method prioritizes DMRs that are controlled by the same mechanism and display coordinated methylation patterns (Supplemental File 3), thereby enriching for loci with interpretable regulatory mechanisms. Notably, DMRs identified through this method are enriched in 5’ UTRs (just downstream of each transcription start site [TSS]) and CTCF-bound regions, supporting their functional relevance in gene regulation (Figure 2E). The quantification of cis- vs. trans-DMRs yielded the same conclusion that cis-regulation outweighs trans-regulation at establishing divergent methylation patterns (Figure 2F).

The causes of cis- and trans-acting divergence in DNA methylation

To identify the molecular mechanisms driving methylation divergence, we analyzed transcription factors (TFs) associated with trans-DMRs and sequence variants associated with cis-DMRs.

TFs represent a major class of trans-acting factors capable of establishing and maintaining methylation patterns across the genome, often by recruiting methyltransferase or demethylase enzymes. Interspecies variation in TF expression levels, binding activity, or target site accessibility can propagate to downstream methylation differences at their binding sites. To identify TFs potentially driving species-specific methylation divergence through such trans-acting mechanisms, we performed TF binding site motif enrichment analysis on trans-DMRs (Figure 3A, Supplemental File 4 and 5)52. This analysis identified several candidate factors, including a pioneer factor FOXA2 (Benjamini-Hochberg corrected FDR = 3.7 × 10⁻⁵ in DA Hu<Ch trans-DMRs) and a chromatin modifier FOXM1 (FDR = 0.002, 4.8 × 10⁻⁶, and 0.021 in CNCC, DA, and HEP Hu<Ch trans-DMRs, respectively), both known to actively reshape DNA methylation landscapes53–60.

Cis- and trans-factors influencing species-specific methylation patterns.

(A) HOMER motif enrichment of trans-DMRs in regions Hu>Ch and Ch>Hu methylated in humans (left panel), and differential expression of corresponding transcription factors in parents (right panel). (B) Fold-enrichments against genomic background of FOXM1, one of the top motifs enriched in Hu>Ch and Ch>Hu trans-DMRs, plotted against Hu and Ch relative gene expression levels in parents. (C) CpG-disrupting SNV and their effects on neighboring CpG methylation within ±25, 50, and 100 bp. The differential methylation patterns around CpG-disrupting SNV variants versus non-CpG-disrupting variants are assessed using two-proportion test (***: p < 0.001, **: p < 0.01, *: p < 0.05) (D) Distances of each cis-DMR vs trans-DMR to the nearest CpG-disrupting SNV site (Mann-Whitney U test; ***: p < 0.001).

If species-specific levels of a TF are responsible for the enrichment of its predicted binding sites in trans-DMRs, then this enrichment may be strongest in cell types with greater species-specific differential expression of the TF. Consistent with this expectation, higher relative expression of FOXM1 and FOXA2 was associated with lower methylation at their predicted binding sites in that same species and cell type. For example, FOXM1 was over 4-fold more highly expressed in human HEPs than in chimpanzee, and its predicted binding sites were highly enriched specifically for trans-DMRs with lower methylation in human HEPs; in contrast, three other cell types with more similar human vs. chimpanzee expression of FOXM1 showed weaker enrichment and little species-specific asymmetry in trans-DMR methylation directionality (Figure 3B). We observed a similar pattern for FOXA2 (Figure 3—figure supplement 1).

Other TFs showed different patterns of association between species-specific expression and trans-DMR methylation. For example, FOXP1 and OTX2 were most differentially expressed in HEPs, consistent with their predicted binding sites also being most enriched in trans-DMRs in this same cell type (Figure 3—figure supplement 1). However, unlike the pioneer factors described above, binding sites for FOXP1 and OTX2 showed similar patterns of enrichment across cell types for both Hu>Ch and Ch>Hu methylated trans-DMRs. This lack of directionality may suggest that these transcription factors are not restricted to either demethylation or methylation, perhaps having a more indirect or context-specific role in regulating the epigenetic landscape61–63, though further work is required to test this hypothesis. Altogether, our trans-DMR analysis nominated many TFs that may contribute to methylation divergence, including several whose trans-DMR binding site motif enrichment is associated with the TF’s differential expression.

To identify the molecular mechanisms driving cis-regulatory methylation differences, we then examined whether local sequence variants could explain cis-acting methylation divergence. CpG dinucleotides are the primary targets of DNA methylation in mammalian genomes, and their presence, density, and spacing are known to influence local methylation patterns64–66. Therefore, single nucleotide variants (SNVs) that either create or disrupt CpG dinucleotides represent a plausible mechanism by which sequence divergence could drive species-specific methylation differences. To investigate this hypothesis, we systematically analyzed SNVs and their associated methylation differences in neighboring CpGs within windows of 25, 50, and 100 base pairs upstream and downstream of each SNV. Our analysis revealed that SNVs affecting CpG dinucleotides—but not other SNVs—are associated with altered methylation patterns of nearby CpG sites up to 50 base pairs from the variant site (Figure 3C). Notably, the effect diminished with increasing distance from the SNV, with the strongest methylation changes observed within 25 base pairs and progressively weaker effects at 50 base pairs. Beyond 50 base pairs, the magnitude of influence of CpG-disrupting variants on neighboring methylation patterns became negligible, suggesting a limited spatial range for these local sequence effects. This distance-dependent relationship indicates that sequence variants affecting CpG density and spacing might influence the local recruitment or activity of DNA methyltransferases or demethylases.

This model predicts that while cis-DMRs should often be close to CpG-disrupting SNVs, trans-DMRs—which arise from regulatory mechanisms independent of nearby sequence variation—should not be. To test this, we compared the genomic distance between each DMR and its nearest CpG-disrupting SNV. Consistent with our model, we found consistently shorter distances for cis-DMRs than trans-DMRs across all cell types examined (Mann-Whitney p < 0.001 for all comparisons)67. This difference suggests that cis-DMRs tend to occur near sequence variants and are likely directly influenced by local changes in CpG content, while trans-DMRs show no preferential association with CpG-disrupting variants. This genomic organization is consistent with trans-DMRs being regulated by diffusible factors such as transcription factors rather than by local sequence variants, supporting a model in which cis-and trans-acting mechanisms operate through fundamentally different molecular pathways to generate species-specific methylation patterns.

Cell type- and species-specific gene expression as a consequence of methylation divergence

Given the well-established repressive effect of promoter methylation on gene expression, we assessed whether methylation differences contribute to expression divergence by analyzing genome-wide correlations between allele-specific promoter methylation and expression levels across different genomic contexts (Figure 4A)68,69. We first examined whether cis-acting variants drive promoter methylation consistently across cell types or in a context-dependent manner. We found that while some promoters are allele-specifically methylated across all cell types, many more showed cell type-specific cis-methylation patterns, suggesting that promoter methylation regulation is often context-dependent (Figure 4B and 4C). However, while DMRs consistently detected across cell types provide strong evidence of shared regulation, DMRs detected in only one cell type may reflect either genuine cell type-specificity or insufficient power to detect them elsewhere. Our estimate of cell type-specific regulation therefore represents an upper bound on the degree of cell type-specific methylation regulation.

Human-chimpanzee allele-specific methylation can be associated with allele-specific expression.

(A) Illustration of the transcriptional consequences of DNA methylation explored in this study. DNA methylation, as a stable cis-acting regulatory mark, often represses transcription. In a hybrid, allele-specific methylation can lead to allele-specific gene expression, although both methylation and its effects on transcription can be context-dependent. (B) We identified many promoters with allele-specific methylation in each cell type, most of which showed allele-specific methylation in only one cell type. (C) Quantification of promoters with pure-cis regulation of methylation across cell types. (D, E, F) Examples of gene expression-methylation correspondence. Methylation tracks show fractional methylation (mC/coverage × 100%) of individual CpGs. Expression values are in TMM (Trimmed Mean of M-values)-normalized CPM (counts per million). (D) CTSF as an example of species-specific, cell type-agnostic DMR, (E) LGALS8 as an example of cell type-specific allele-specific methylation, and (F) HOXA9 as an example of cell type-specific but species-agnostic methylation patterns. (G) Genome-wide profile of expression-methylation correlations (binned by distance of methylated region to the TSS) for genes with both promoter methylation and expression under cis-regulation (dark red and dark blue) or trans-regulation (light brown and light gray). Detailed statistics are provided in Supplemental File 6. Figure 4A was created using BioRender, and is published under a CC BY-NC-ND 4.0 license.

Several specific examples illustrate different patterns of methylation-expression coordination (Supplemental Files 5 and 6; see Methods). The promoter of CTSF is a cell-type agnostic cis-DMR with consistent Hu>Ch methylation across cell types corresponding to coordinated Ch>Hu expression (Figure 4D and Figure 4—figure supplement 1). In contrast, a putative enhancer in LGALS8 represents a case where methylation divergence is confined to specific cellular contexts. The candidate enhancer overlaps with a cis-DMR with Hu>Ch methylation and LGALS8 exhibits allele-specific Ch>Hu expression exclusively in CNCCs, while having no divergence in methylation or expression in other cell types (Figure 4E and Figure 4—figure supplement 2). A third type of pattern is exemplified by the HOXA9 promoter, with cell type-specific hypomethylation corresponding to cell type-specific expression only in SKM (Figure 4F). These examples illustrate how different regulatory mechanisms can produce distinct patterns of methylation-expression coordination and how epigenetic changes may contribute to both species-specific and cell type-specific regulatory evolution.

Previous studies have reported weak but significant correlations between DNA methylation and gene expression divergence in primates7,12,14,16,70. Consistent with these findings, our genome-wide estimates of human–chimpanzee transcriptome divergence produced similar weak but significant correlations with methylation divergence (Figure 4—figure supplement 3 and Supplemental File 6). Our comparison of CpG island (CGI) vs. non-CGI promoters also yielded trends consistent with past studies showing that CGI promoter methylation correlates more strongly with gene expression repression than does non-CGI promoter methylation71–73 (Figure 4—figure supplement 4).

However, such genome-wide correlations may mask stronger relationships that occur within specific regulatory contexts where methylation and expression are more directly coupled. We hypothesized that our hybrid system could help identify genes exhibiting stronger methylation–expression coupling by stratifying them based on regulatory mechanisms. We reasoned that genes whose methylation and expression are both influenced by cis-acting variants— potentially driven by the same underlying genetic differences— should display tighter coupling than genes regulated through independent mechanisms.

To test this, we evaluated the extent to which methylation and expression show a negative relationship (i.e., higher methylation associated with lower expression) consistent with a repressive effect of methylation. In both CNCC and DA hybrid cells, we found that genes with cis-regulated expression were more likely to exhibit repressive patterns than those with trans-regulated expression (Figure 4—figure supplement 5). This is consistent with DNA methylation acting as an upstream cis-regulatory mechanism: genes with stronger cis-regulation are expected to show tighter methylation-expression coupling due to the greater contribution of local regulatory factors, including methylation itself. Further restricting the analysis to genes with concordant regulatory mechanisms revealed that genes with both methylation and expression being cis-regulated (“cis–cis” genes) exhibited significantly stronger negative methylation–expression correlations than “trans–trans” genes (Figure 4G, see Supplemental File 6 for detailed statistics). This supports the hypothesis that methylation–expression coupling is strongest when both are driven by local genetic variation, perhaps through shared causal variants.

The sign test identifies lineage-specific selection on DNA methylation between human and chimpanzee

To identify pathways that have evolved under lineage-specific selection, we used a two-step binomial sign test approach that tests for non-random directional bias in methylation-expression patterns in hybrids1,24,25,27,74–78. The use of hybrid cell lines is critical for this analysis as it ensures that different promoters represent independent observations, controlling for trans-acting factors and environmental variation that could otherwise confound the detection of cis-regulatory effects. This test operates on the logic that neutral evolution should act like flipping a fair coin: if human-biased and chimpanzee-biased changes occur at equal background frequencies, then methylation and expression changes of genes within a pathway would accumulate randomly, showing no preference for human-biased versus chimpanzee-biased directionality. In contrast, lineage-specific selection would bias the coin flip, potentially causing functionally related gene sets to consistently favor one direction over the other rather than splitting evenly by chance.

To maximize our ability to find phenotypically impactful divergence, we adapted the standard sign test to account for directionality of both methylation and gene expression. In the first step, we applied a standard binomial sign test to all promoters showing ASM, counting the number of genes showing human-biased versus chimpanzee-biased methylation changes within each gene set. The background probability for this test was defined as the overall fraction of promoter ASM in each cell type that was human-biased. This initial analysis identified gene sets showing significant enrichment for directional methylation changes, establishing candidates for lineage-specific regulatory divergence. In the second step, we restricted our attention to genes showing a negative association between ASM and ASE. This subset represents cases where methylation changes are most likely to have a simple and direct repressive effect on transcription, filtering out more complex regulatory scenarios where methylation and expression changes may be uncoupled or subject to competing regulatory influences.

Gene sets showing significant deviation from neutral expectation in the ASM/ASE concordance test (nominal binomial p-value < 0.05 and FDR < 0.25 after multiple test correction; see Methods), with supporting evidence of directional bias in the methylation-only sign test (nominal binomial p-value < 0.05), are interpreted as likely targets of lineage-specific selection on promoter methylation that could directly impact gene expression, establishing a potential mechanistic connection to phenotype (Figure 5, Supplemental File 7). Compared to a standard sign test that focuses solely on directionally biased gene expression, this more stringent two-step filtering approach may identify higher-confidence candidates where regulatory divergence is reflected at both the methylation and expression levels74.

Gene sets with evidence of lineage-specific selection on methylation and gene expression.

The length of the bars indicates the number of genes in each gene set with allele-specific expression or methylation in each direction (human or chimpanzee-biased). The bars in darker colors represent the number of genes where higher methylation is associated with repressed gene expression, whereas lighter colors represent genes where higher methylation is associated with increased gene expression.

Among gene sets from Kyoto Encyclopedia of Genes and Genomes (KEGG) and Human Phenotype Ontology (HPO)79,80, we identified several with clear relevance to human-specific traits. In iPSC hybrids, multiple gene sets including “poor speech,” “highly arched eyebrow,” and “growth delay” all showed Hu>Ch methylation and Ch>Hu expression (Figure 5, Figure 6A and Figure 6B). In the SKM hybrids, genes related to “inability to walk” showed significant Hu>Ch methylation and Hu<Ch expression (Figure 5). In DA hybrids, genes involved in “intellectual disability” show significant Hu<Ch methylation and Hu>Ch expression (Figure 5 and Figure 6C). In HEP hybrids, we identified the “Hepatitis C” gene set with significant Hu>Ch methylation and Hu<Ch expression (Figure 5 and Figure 6D). In primary DPSC samples, we identified significant bias toward Hu<Ch methylation and Hu>Ch expression in genes related to tooth eruption timing, consistent with known differences in dental development between humans and chimpanzees, where human permanent molars emerge approximately 6 years apart compared to 3-4 years in chimpanzees (Figure 5—figure supplement 1)81–83. However, the DPSC sign test results should not be interpreted as evidence of lineage-specific selection due to the use of parental rather than hybrid cell lines, which may introduce dependencies among observations due to trans-acting effects, potentially inflating statistical significance.

Directional bias in allele-specific methylation and gene expression identifies candidate genes for human-specific phenotypes.

Allele-specific expression (ASE) is plotted against allele-specific methylation (ASM). Orange dots represent genes consistent with repressive methylation (e.g. Hu>Ch ASE and Hu<Ch ASM) and purple dots represent genes consistent with activating methylation (e.g. Hu>Ch ASE and Hu>Ch ASM). Bar plots represent counts of genes with repressive methylation that exhibit expression changes in directions either consistent with (light green) or opposite to (light pink) what would be expected based on phenotypic differences between humans and chimpanzees. A) The “highly arched eyebrow” gene set in iPSC hybrids shows consistent Hu<Ch ASE and Hu>Ch ASM for genes with repressive methylation. B) The “growth delay” gene set in iPSC hybrids shows significant bias towards Hu<Ch ASE and Hu>Ch ASM for genes with repressive methylation. C) The “intellectual disability, moderate” gene set in DA shows consistent Hu>Ch ASE and Hu<Ch ASM for genes with repressive methylation, with genes including GRIK2, TUBB3, EGF and CC2D2A being among the genes with highest magnitude ASE and ASM. D) The “Hepatitis C” gene set in HEP shows consistent Hu<Ch ASE and Hu>Ch ASM for genes with repressive methylation. D) The “poor speech” gene set in iPSC hybrids represents a potentially compensatory pathway where most genes with repressive methylation show Hu<Ch ASE and Hu>Ch ASM, whereas the genes with highest magnitude of ASE and ASM (TUBB3 and GRIK2) show Hu>Ch ASE and Hu<Ch ASM.

We then analyzed individual genes within each candidate gene set to assess whether expression bias in associated phenotypes aligns with known human-chimpanzee trait differences, focusing on genes showing repressive methylation. Multiple gene sets demonstrated directional bias in gene expression consistent with the direction of phenotypic divergence found between humans and chimpanzees. For instance, in the “highly arched eyebrow” gene set—where most genes show human down-regulation in iPSC hybrids—COLEC11 (which has 8-fold lower expression from the human allele) stands out as an outlier. COLEC11 acts as a guidance cue in directing neural crest cell migration during early development, and in humans its loss-of-function (LOF) causes craniofacial dysmorphism including highly arched eyebrows and underdeveloped brow ridge, both of which are traits unique to humans (Figure 5 and Figure 6A)84. In the “growth delay” pathway, most genes with repressive methylation (whose LOF in humans and other species leads to delayed growth)62–65 show lower expression from their human alleles, consistent with humans’ substantially delayed growth pattern compared to chimpanzees (Figure 6B)85,86. Many of the genes in DA whose LOF causes “intellectual disability” may have contributed to the evolution of human cognitive abilities. For example, the high-magnitude genes with Hu>Ch ASE and Hu<Ch ASM such as GRIK2, TUBB3, EGF and CC2D2A are essential for excitatory synaptic plasticity and long-range axon wiring (Figure 6C)87–91. In addition, most genes whose inhibition or LOF increases susceptibility or severity of Hepatitis C (e.g. MX1 and OAS3) show lower expression from human alleles in HEP, consistent with human’s heightened vulnerability to Hepatitis C viral infections (Figure 6D)92–94.

Interestingly, we also discovered gene sets where the direction of most gene expression differences predicted the opposite of a known human/chimpanzee phenotypic difference. For example, in the “poor speech” gene set, where loss-of-function is associated with impaired speech in humans, we counterintuitively observed that most genes were down-regulated in humans compared to chimpanzees. However, the two genes with the largest fold-changes, TUBB3 and GRIK2, showed higher expression from the human allele (Figure 6E). Notably, these two genes are also among the large-magnitude genes in the predominantly human up-regulated “intellectual disability” pathway, which does align with species-specific phenotype (Figure 6C). This suggests the speculative possibility of compensatory evolution in which large, pleiotropic effects of a few genes are counterbalanced by smaller changes in many genes (Figure 6C and Figure 6E; also see Discussion). Our previous study found that genes affecting voice box show particularly extensive down-regulation in humans, consistent with our observation in the “poor speech” pathway25.

Collectively, these results provide the first evidence for lineage-specific selection on human-specific DNA methylation. Our use of hybrid cell lines was critical to this discovery, as it allowed us to isolate cis-acting regulatory changes from confounding trans-acting and environmental effects. This approach reveals that polygenic selection has acted on entire sets of functionally related genes whose expression is modulated through DNA methylation. Importantly, by connecting lineage-specific methylation divergence to expression changes within functionally coherent pathways, we can generate plausible hypotheses linking regulatory evolution to human-specific phenotypes. These include craniofacial morphology, dental development, viral infection susceptibility, and neurodevelopmental characteristics underlying speech and cognition (Figures 5 and 6, Figure 5—figure supplement 1).

Discussion

Our investigation using human-chimpanzee hybrid tetraploid cells revealed that cis-regulatory mechanisms are the primary drivers of genome-wide methylation divergence between humans and chimpanzees across multiple cell types, including dopaminergic neurons, skeletal muscle, iPSCs, hepatocytes, and cranial neural crest cells. This finding aligns with previous studies highlighting the predominant role of cis-regulatory elements in evolutionary adaptations, as they enable precise, context-specific gene regulation while minimizing the potentially deleterious pleiotropic effects of trans-acting factors6. We demonstrated that most differentially methylated sites arise through cis-regulatory mechanisms, with trans-regulation and cis-trans interactions playing secondary roles. We identified clusters of CpG sites with consistent regulatory mechanisms, forming coherent differentially methylated regions (DMRs). These DMRs predominantly overlap with functionally relevant genomic features including 5’ UTRs and upstream promoter regions, underscoring their regulatory significance.

Our mechanistic analysis revealed distinct pathways through which cis- and trans-regulatory factors establish species-specific methylation patterns. For cis-regulation, we identified CpG-altering sequence variants as a direct mechanism linking genomic divergence to epigenetic differences. These variants influence methylation patterns of neighboring CpG sites with effects diminishing with genomic distance. Importantly, cis-DMR clusters are significantly closer to CpG-disrupting SNVs than trans-DMRs across all cell types, providing strong evidence that local sequence variation directly shapes cis-regulatory methylation patterns, perhaps through altered CpG density affecting recruitment of DNA methyltransferases or demethylases.

While we identified substantial numbers of allele-specifically methylated promoters in all cell types, the majority showed cell type-specific patterns. Our examples illustrate this diversity: the CTSF promoter exhibits consistent cross-cell-type cis-methylation divergence (Figure 4—figure supplement 1) with corresponding expression changes; an enhancer near LGALS8 shows cell type-specific allele-specific methylation and overlaps with a homogeneous cis-DMR exclusively in cranial neural crest cells; and the HOXA9 promoter displays cell type-specific methylation patterns with corresponding context-dependent expression effects. These findings highlight how the broader regulatory context of each cell type impacts the effects of cis-regulatory variants.

Our genome-wide correlation analyses confirmed the well-known pattern that while promoter methylation generally shows statistically significant inverse relationships with gene expression, these correlations are weak when examined across all genes. However, by leveraging our hybrid cell system to stratify genes based on regulatory mechanism, we uncovered significantly stronger negative correlations between methylation and expression when both are cis-regulated compared to genome-wide estimates or trans-regulated genes. This finding suggests that the repressive effects of species-specific methylation are strongest when both layers of regulation are driven by the same or closely linked genetic determinants.

Our two-step binomial sign test approach in hybrids identified multiple gene sets showing coordinated directional shifts in methylation and expression that significantly exceed neutral expectations. Many of the pathways we identified are related to known human-chimpanzee phenotypic differences: intellectual disability in hybrid DAs, growth delay and highly arched eyebrow in hybrid iPSCs, hepatitis C in hybrid HEPs, and tooth eruption timing in parental DPSCs (Figure 5, 6, Figure 5—figure supplement 1). These findings represent the first evidence for lineage-specific selection acting on human-specific DNA methylation. By restricting analysis to genes showing negative methylation-expression associations—where methylation most likely exerts direct transcriptional repression—we identified high-confidence candidates where epigenetic divergence has likely functional consequences. Our results suggest that cis-regulatory changes in DNA methylation represent an important substrate for adaptive evolution, complementing protein-coding and other cis-regulatory changes in generating human-specific traits.

We developed several methodological innovations tailored to the unique challenges of comparative epigenomics. First, our beta-binomial framework accounts for the count-based nature of bisulfite sequencing data and explicitly models allele-specific patterns in hybrid versus parental contexts. Second, our changepoint detection algorithm identifies coherent DMRs with minimal regulatory heterogeneity, prioritizing functionally organized regulatory units over isolated CpG sites. Third, the two-step sign test we introduce provides a method to identify lineage-specific selection that links methylation and corresponding changes in gene expression. These methods can be readily applied to other taxa where hybrid organisms can be generated.

However, our study also has important limitations. First, while hybrid cells provide powerful controls for trans-acting variation, in vitro differentiation protocols do not fully recapitulate the complexity and temporal dynamics of in vivo development. Therefore, cell lines are limited in their ability to capture species-specific methylation or expression profiles that are specific to a certain developmental stage or in vivo context. Second, pathway analysis of DPSCs used primary human and chimpanzee samples rather than hybrids, with trans-acting factors potentially inflating statistical significance in the sign test; these results should therefore be interpreted not as evidence for positive selection, but rather as an interesting pattern that could have important connections to phenotypic differences between humans and chimps. Third, while we identified strong associations between transcription factor binding motifs and trans-DMRs, definitive demonstration of causality requires targeted experimental perturbations. Fourth, our study examined a limited number of cell types, and methylation divergence in most tissues remains unexplored.

Comparative gene expression analyses between humans and closely related species offer a powerful way to explore the molecular basis of species-specific traits and disease susceptibilities. However, it remains challenging to infer phenotypic differences through the lens of human clinical ontologies. To start with, HPO terms are largely derived from pathological human data and predominantly capture phenotypic consequences of loss-of-function mutations in a clinical context. When comparing expression levels between species, one might assume that lower expression corresponds to a loss-of-function-like phenotype. However, the relationship between expression dosage and function is typically nonlinear95, and complete gene knockouts in a model organism or loss of function mutations in patients are not equivalent to moderate downregulation. This fact may help explain cases where our sign test results suggest the opposite directionality as compared to human/chimpanzee phenotypic differences (Figure 6E). Alternatively, humans and chimpanzees may achieve a similar phenotype through distinct configurations of regulatory networks. For example, an adaptive down-regulation of a particular gene may be compensated for by up-regulation of other genes in the same pathway, maintaining overall pathway function. In other words, compensatory changes may complicate the relationship between gene expression and phenotype, as may be the case for the “poor speech” gene set (Figure 6E)24.

Several important directions emerge from this work. First, extending this framework to additional cell types and developmental time points would reveal whether the predominance of cis-regulation we observe is universal or varies across developmental contexts. Longitudinal sampling during differentiation could identify critical windows when methylation patterns are established and reveal whether human-chimpanzee differences emerge gradually or through discrete developmental transitions. Second, integrating chromatin accessibility (ATAC-seq) and histone modification (ChIP-seq) data would provide a more complete picture of the regulatory landscape, clarifying how DNA methylation interacts with other chromatin features to establish gene expression patterns. Third, direct examination of transcription factor binding using techniques like CUT&RUN in both species could validate the predicted relationships between transcription factor occupancy and methylation patterns at trans-DMRs. Finally, extending these analyses to three-way comparisons including other great apes would provide phylogenetic resolution to distinguish human-specific changes from chimpanzee-derived changes.

Collectively, our findings provide a framework for understanding how DNA methylation divergence contributes to human-specific traits. We demonstrate that while both cis- and trans-regulatory mechanisms shape interspecies methylation differences, cis-acting factors predominate, thus directly linking genomic sequence variation and epigenetic divergence. The coordinated nature of many independent methylation-expression changes across phenotypically relevant pathways suggests that epigenetic modifications represent important targets of natural selection in human evolution, contributing to the regulatory innovations underlying uniquely human phenotypes.

Materials and Methods

Cell culture and differentiation

We utilized two previously characterized human-chimpanzee hybrid iPSC lines generated through in vitro cell fusion maintained on matrigel in mTeSR1 or mTeSR Plus medium (Stem Cell Technologies cat #85850 or cat #100–0276)24. The Columbia Stem Cell Core Facility differentiated the iPSCs into multiple cell types including dopaminergic neurons (DA), skeletal myocytes (SKM), and hepatocytes (HEP) using published protocols27,46,96–104. Cranial neural crest cells were differentiated as described25. Primary dental pulp stem cells (DPSC) samples were obtained from humans and chimpanzees as described25.

RNA-sequencing and allele-specific expression analysis

All samples were cryopreserved in liquid nitrogen before RNA extraction105. Cells were gently thawed and then washed with PBS and cell pellets were collected via centrifugation at 1000 RPM for 5 min. RNA extraction was performed using the RNeasy Mini Kit (Qiagen, 74104) and DNase digestion was performed using RNase-Free DNase Set (Qiagen, 79254) following standard protocols. RNA quality was assessed using the Agilent Bioanalyzer RNA Pico assay. Only samples with an RNA integrity number (RIN) greater than or equal to 7 were used to prepare cDNA libraries. Libraries were prepared using TruSeq Stranded mRNA kits (Illumina, 20020594) and sequenced on Illumina HiSeq 4000 to generate paired-end reads. An allele-specific expression pipeline adapted from our previous work was implemented, using dual reference genome alignment to hg38 and panTro6 to eliminate mapping bias24,25. Reads were processed with SeqPrep, mapped using STAR, deduplicated with Picard, and corrected for allelic bias using Hornet. DESeq2 was used for differential expression analysis with FDR correction, defining significant allele-specific expression as FDR < 0.05 with consistent results across both reference genomes24,25,27,106,107. For details, see “Generation of allele-specific count tables” and “Identifying genes with ASE” in the Methods section in Wang et al. (2024)27.

Bisulfite-seq library preparation, sequencing and mapping

DNA was extracted and processed, and whole-genome bisulfite sequencing (WGBS) and reduced representation bisulfite sequencing (RRBS) was performed at Admera Health Biopharma Services using Illumina Novaseq 6000. To minimize reference mapping bias in cross-species comparisons, reference genomes were first masked at previously identified human-chimpanzee SNV positions using bedtools maskfasta, reducing the potential for allelic mapping preferences27,108. Reference genome preparation was performed using Bismark’s bismark_genome_preparation with Bowtie2 as the aligner, which converts unmethylated cytosines in silico to enable bisulfite-aware alignment109,110. All samples were aligned to both masked human (hg38) and chimpanzee (pantro6) reference genomes to assess mapping quality and species-specific biases. Reads were quality-trimmed and adapter sequences removed using Cutadapt and Trim Galore with default parameters for RRBS libraries, including the removal of artificially filled-in cytosines from the 3’ end of reads111,112. Trimmed reads were aligned using Bismark, and methylation levels were calculated as the ratio of unconverted cytosines to total coverage at each CpG site109. For hybrid samples, SNPsplit was used to assign reads to either human or chimpanzee allele113. Strand information was collapsed for symmetric CpG sites and genomic annotations were lifted between genomic assemblies hg38 and pantro6 using UCSC LiftOver with reciprocal mapping to ensure orthologous regions114,115.

PCA and hierarchical clustering

Fractional methylation of individual CpG sites was estimated using bismark_methylation_extractor from Bismark109. CpG sites shared across samples were identified using bedtools intersect108. PCA was performed using PCASamples from methylKit across all samples and samples within each cell type116. Euclidean distance matrices were computed across all samples for hierarchical clustering analysis, with visualization performed using the pheatmap package117. Clustering patterns were assessed to validate cell type identity and characterize species-specific methylation signatures.

Beta-binomial regulatory classification framework

Methylation data from human (H) and chimpanzee (C) parental and hybrid samples were analyzed using a beta-binomial generalized linear modeling framework to account for overdispersion inherent to bisulfite sequencing count data. The data included measurements for parental H-genome (ParH), parental C-genome (ParC), hybrid H-allele (HybH), and hybrid C-allele (HybC) for each genomic locus. For each genomic locus (site or annotated region), methylated counts were modeled using two complementary beta-binomial GLM models to separately evaluate cis and trans regulatory effects:

Model I: Cis effect test with methylation data on hybrid alleles

Model II: Trans effect test using System x Allele interaction term with methylation data in both hybrid and parental alleles

Where π is the probability of methylation, System distinguishes parental from hybrid samples, Allele distinguishes between human and chimpanzee methylation, and the interaction term captures any signals involving the change of allelic differences across generations. Statistical significance was assessed through likelihood ratio tests comparing nested models, testing for system effects, allele effects (cis-regulation), and system-by-allele interactions (trans-regulation). Genomic loci were classified into five categories: conserved, trans-only, cis-only, cis x trans interaction, and cis + trans effects:

Conserved: No significant allelic differences in either generation (neither parental nor hybrid allelic tests significant)

Trans only: Significant allelic difference in parents but not hybrids, indicating trans-acting regulation

Cis only: Significant allelic differences in both systems with no significant interaction, indicating consistent cis-acting effects

Cis × trans: Significant allelic differences in both systems with significant interaction, where effects occur in opposite directions

Cis + trans: Significant allelic differences in both systems with significant interaction, where effects are in the same direction

Regional analyses combined CpG-level methylation across defined genomic intervals (for example, promoters). Regional methylation was quantified as pooled fractional methylation, calculated as the total number of methylated cytosines divided by the total coverage across all CpG sites within each region. Nominal p-values < 0.05 from both models were used to provide a schematic overview of site-level CpG methylation in scatterplots and bar plots, preserving the full dynamic range of the data and providing input features for downstream de novo DMR identification. For reporting significant candidate loci under each regulation category as well as formal inference of region-level regulation, Benjamini–Hochberg false discovery rate (FDR) was calculated to account for multiple testing and to evaluate the robustness of inferred relative contribution of regulatory mechanisms across significance thresholds60. The same classification and visualization were employed using FDR-corrected p-values on loci and regions to confirm that the relative contributions of regulatory classifications were not driven by significance cutoffs alone (Figure 2—figure supplement 1 and 2). To assess the robustness of cis-over-trans contributions to methylation with respect to the significance cutoffs, the proportion of significant loci and regions assigned to each regulatory category was estimated across a range of FDR cutoffs (FDR < 0.05, 0.1, 0.2, 0.25), confirming that the observed cis dominance over other divergent regulation categories (including trans) was robust to the choice of cutoff (Figure 2—figure supplement 1 and 2).

Changepoint detection for identifying differentially methylated regions with homogeneous cis- or trans-acting methylation divergence

To identify clusters of CpG sites in the same regulation category, we employed the Pruned Exact Linear Time (PELT) algorithm implemented in the changepoint R package to detect shifts in mean methylation scores within each chromosome segment while accounting for a manually specified penalty against over-fitting118–120. We adapted this method to identify regions in the genome where neighboring CpG sites are consistently classified by the previously mentioned beta-binomial model to fall under the same regulatory classification. We first segmented the entire genome into large regions where a break is assigned whenever neighboring CpG sites are more than 5000 base pairs apart. Then, we transform all CpG sites on a segment into binary sequences, where the focal regulation category is 1 and alternative categories are 0. To identify trans-DMRs, we set focal pure trans regulation to 1, and all other regulation categories to 0. To identify cis-DMRs, we set focal pure cis regulation to 1, and all other regulation categories to 0. To identify trans-DMR specifically for HOMER enrichment inputs, we used a relaxed gradient scoring approach. We set focal pure trans regulation to 1, conserved to 0.75, additive (cis + trans) and interactive (cis x trans) regulation to 0.5 and pure cis regulation category to 0. To determine the optimal penalty parameter, we applied the CROPS (Changepoints for a Range Of Penalties) method, testing penalty values ranging from 0.01 to 2.0. For each penalty range giving the same number of segments, we recorded the number of detected changepoints and corresponding model fit statistics. The optimal penalty for each segment was determined using elbow point detection via the R package kneedle, which identifies the point of maximum distance from line connecting endpoints in the penalty-versus-changepoints curve121. We then applied a homogeneity filter requiring segments to have a minimum target proportion of 0.6, where the target proportion is the fraction of CpG sites within the segment that (i) belong to the target regulatory class under the beta-binomial cis/trans classification and (ii) share a consistent species bias in methylation (all human-biased or all chimpanzee-biased), to ensure biological relevance of identified regions. For segments where fewer than three distinct penalty values yielded different numbers of changepoints, we selected the penalty that maximized the number of segments passing the homogeneity threshold. Segments without detected changepoints were evaluated directly against the homogeneity criterion to assess their potential as single-segment DMRs.

To refine segment boundaries and remove spurious edge effects before merging overlapping DMRs, we implemented a post-processing trimming step that removes segment edges containing sites with regulatory patterns inconsistent with the dominant signal within each segment. This ensures that identified DMRs represent coherent regions of coordinated methylation changes rather than artifacts of the segmentation process.

Identification of cell-type specific DMRs, species-specific DMRs, and cell-type specific species-specific DMRs

Examples of cell-type specific DMRs (HOXA9), species-specific DMRs (CTSF), and cell type-specific species-specific DMRs (LGALS8) were identified in CNCC, DA, iPSC, SKM and HEP hybrid samples using the DSS-single (Dispersion Shrinkage for Sequencing data)122–125. For cell type-specific DMR identification, we used the formula ∼species + cell type + species:cell type to test for cell type-specific methylation differences, while species-specific DMRs analysis employed ∼species to identify consistent species differences across cell types. Downstream identification of cell-type specific species-specific DMRs identifies the regions where the interaction term species:cell type is significant. The DSS analysis incorporated smoothing across neighboring CpG sites with a smoothing span of 500 bp and required a minimum of 5 CpG sites per region. Statistical significance was determined using Wald tests with FDR < 0.05, and effect size filtering retained regions with methylation difference ≥20%. Integration with gene expression data was performed through Wilcoxon rank-sum tests to identify cell type-specific gene expression patterns associated with cell-type specific DMRs, while permutations (n=1000) were used to assess the significance of coordinated methylation-expression changes in species-specific patterns, which accounts for any background correlation between neighboring genomic features.

Expression-methylation correlation analysis

Human transcript coordinates and gene body annotations were obtained from the hg38.knownGene.gtf.gz file via UCSC Table Browser114, with transcript start positions designated as transcription start sites (TSS). CpG methylation data aggregated within ENCODE-defined promoter regions were analyzed using a beta-binomial classification framework to determine regulatory categories, while corresponding gene expression data underwent parallel hypothesis testing using a negative-binomial model (ENCODE Project Consortium, et al., 2020). Genes with the same classification for both methylation and expression—either both cis or both trans—were then analyzed separately. For genes within each regulatory category, the 2 kb surrounding each TSS was split into 10 bins of 200 bp where individual CpG sites were pooled and averaged as fractional methylation values and correlated with expression TPM values. Custom scripts used for binning and correlation are adapted from beta_profile_region.py from CpGtools (Wei et al., 2021).

Sign test selection analysis

To identify coordinated methylation-expression changes potentially under selection, we used a binomial sign test framework that accounts for background probabilities of concordant changes24,25,27,31,128,129. Promoter regions were obtained from the ENCODE Project Consortium126. Canonical transcripts were downloaded from the UCSC Table Browser hg38 assembly. For genes with multiple annotated promoters, we selected the promoter that is the closest to 5’ UTR of the gene, to ensure each gene is assigned one corresponding promoter114. Promoter regions were mapped to canonical transcripts, and functional enrichment using pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) and Human Phenotype Ontology (HPO) was assessed using directional sign tests that leverage the independence of cis-regulatory regions to detect non-random patterns indicative of lineage-specific selection rather than neutral drift1,24,25,27,75–79. For the methylation-only sign test, gene sets with more than 200 genes or fewer than 20 genes with corresponding promoter methylation data were excluded from the analysis. Lineage-specific selection on a gene set was inferred if there was a statistically significant directionality bias in chimpanzee vs. human promoter ASM. For sign test on genes with putatively repressive methylation—that is, genes where increased methylation within 1 kb of their TSS correspond to decreased expression—we first classified genes as either consistent with repressive methylation (either Ch>Hu methylation paired with Hu>Ch expression, or Hu>Ch methylation paired with Ch>Hu expression) or not consistent. Gene sets with more than 200 genes or fewer than 10 genes with corresponding methylation and expression data were excluded from the analysis. We then reran the sign test using a new background probability defined as the overall fraction of genes with repressive methylation in each cell type. FDRs of the resulting binomial p-values were estimated by randomly permuting the signs of all genes and performing the binomial sign test 104 times. To calculate FDR, we compared each pathway’s empirical binomial p-value against the distribution of permuted p-values. The expected number of false positives was estimated as the average number of pathways reaching that significance level per permutation. FDR was then defined as this expected count of false positives divided by the observed count of pathways with empirical p-values at least as extreme.

cis-DMR sequence variant analysis and trans-DMR motif enrichment analysis

High-confidence human-chimpanzee single nucleotide variants (SNVs) were identified using a dual-reference approach with stringent filtering criteria as described previously27. Allele-specific methylation within different genomic ranges (25bp, 50bp, 100bp) of CpG-disrupting variants was calculated. Proportions of regions with Hu>Ch and Hu<Ch methylation were calculated for SNVs in three categories: SNVs that disrupt CpG dinucleotides in humans, SNVs that disrupt CpG sites in chimpanzees, and SNVs that lead to no interspecies changes in CpG content between the species. The significance of the difference in relative proportions of Hu>Ch and Hu<Ch of SNVs in the three categories was assessed using a two-proportion test. For proximity tests, the Mann-Whitney U test67 was used to compare the distribution of distances from all cis or all trans-DMRs to the closest CpG-disrupting SNV. HOMER’s findMotifsGenome.pl was used to test for motif enrichment in the trans-DMR set52.

eLife Assessment

This study presents an important examination of the role of cis-acting versus trans-acting genetic variation on DNA methylation divergence between humans and chimpanzees, including its consequences for gene expression. By differentiating fused interspecies tetraploid cell lines into multiple cell types, the study provides compelling evidence for the importance of cis-acting changes, but incomplete evidence that these changes are of importance for adaptive trait evolution in humans. This work will be of interest to biologists and evolutionary anthropologists studying the evolution and genetics of gene regulation, particularly in primates.

Reviewer #1 (Public review):

Ma et al. use human-chimpanzee tetraploid cells across different cell types to identify the genetic causes and then transcriptomic consequences of divergence in DNA methylation. They conclude that the evolution of DNA methylation is driven primarily by cis-regulatory changes, and that the evolution of CpG sites contributes to cis-regulation, while transcription factor expression underlies some trans changes. They then argue that divergence in DNA methylation is associated with changes in gene expression and may contribute to human phenotypes.

The tetraploid model is able to provide compelling evidence that most regulatory evolution occurs due to cis-regulatory changes. My only concern is that the extent of trans-changes may be overstated, as almost all are eliminated by changing from a nominal p-value criterion to even a 25% false discovery rate. The follow-up analyses are incomplete with major gaps. The authors focus on single potential mechanisms for cis- and trans-changes, but it is not clear to what degree these mechanisms explain the extent of cis and trans changes. There are also other mechanisms which are not investigated, such as the importance of TF binding sites for cis-regulatory evolution. While likely beyond the scope of this work, communicating these areas for future work would have helped define the niche for this manuscript.

Next, the authors seek to show that differences in DNA methylation are functionally relevant. Consistent with previous results, they show that differences in DNA methylation are (weakly) associated with changes in gene expression. They hypothesize that genes with concordant regulatory elements should exhibit greater methylation-expression coupling than other genes and show that cis-expression/cis-methylation pairs are more strongly correlated than trans/trans pairs. However, I worry that this result could be confounded by larger effect sizes for cis-changes than trans effects. I also think that looking at cis/trans or trans/cis changes would have been useful to directly test the driving hypothesis. Another limitation is that this analysis is limited to promoter regions. It is not clear how many divergent DMRs are included and how many of those genes have differences in expression. The key question is whether differences in DNA methylation are functionally important, and the answer provided by these analyses is "sometimes".

Finally, the authors make a case for lineage-specific selection on DNA methylation that is connected to human traits. This evidence was not convincing. In fact, it is even said that these tests cannot be interpreted as evidence of lineage-specific selection (lines 399-401), so I am confused why these results are framed as testing for selection. The evidence better supports an argument connecting DNA methylation to human phenotypes.

In conclusion, I think this study provides a valuable resource for differences in DNA methylation between humans and chimpanzees across tissues, and provides important insight into the relative abundance of cis and trans regulatory evolution. Additional research is necessary to investigate the underlying regulatory mechanisms, and more care needs to be taken in exploring the functional consequences.

Reviewer #2 (Public review):

Summary:

This manuscript investigates the causes and consequences of human-specific DNA methylation divergence relative to chimpanzees. The main aim of this study is to disentangle cis- and trans-regulatory contributions to DNA methylation differences, which the authors address using an innovative interspecies hybrid cell system differentiated into multiple cell types. This design allows them to control for trans-acting environments and directly compare allelic regulation.

The authors show that cis-regulatory mechanisms dominate DNA methylation divergence and that methylation-expression coupling is strongest when both are cis-regulated. They further explore potential mechanisms underlying these patterns, including CpG-disrupting mutations and transcription factor-associated trans effects, and identify pathways that may reflect lineage-specific regulatory evolution.

This study provides a valuable dataset and a compelling framework for understanding how local sequence variation contributes to epigenetic and transcriptional divergence, with likely broad impact in comparative and evolutionary genomics.

Strengths:

A major strength of this study is the use of human-chimpanzee hybrid cells, which provides a powerful system to disentangle cis- and trans-regulatory effects in a shared cellular environment. This experimental design allows for a more definitive assessment of regulatory mechanisms than traditional cross-species comparisons.

The study also benefits from the inclusion of multiple differentiated cell types, increasing the robustness and generality of the conclusions. The consistent observation that cis-regulatory mechanisms dominate methylation divergence across these contexts is well supported by both CpG-level and DMR-level analyses.

Another important contribution is the finding that methylation-expression coupling is strongest when both are cis-regulated. This provides a mechanistic explanation for previously observed weak global correlations between methylation and gene expression. Given that the nature of regulatory evolution is likely highly heterogeneous, this study adds valuable insights and guidelines for future investigations. I recommend that the authors provide a list of cis-cis-regulated promoters and their associated genes, which would be a valuable resource for the field.

The application of the two-step sign test identifies biologically relevant pathways, suggesting links between epigenetic divergence and human-specific traits.

The dataset itself, namely, comprehensive DNA methylation and gene expression across multiple cell types in shared cellular contexts, as well as a primary cell type, is a valuable resource for the field. Additionally, the application of the two-step sign test identifies biologically relevant pathways, suggesting links between epigenetic divergence and human-specific traits.

Weaknesses:

Although the authors identify transcription factors associated with differential methylation, it is unclear what proportion of differentially methylated CpGs or DMRs can be attributed to these factors. Providing a quantitative estimate would help assess the relative contribution of trans-acting regulation.

The analysis of CpG-disrupting mutations is interesting but raises two concerns. First, other classes of variants-such as transcription factor binding site-disrupting mutations-could also influence local methylation patterns and are not considered here. Second, the causal direction remains ambiguous: CpG-disrupting mutations may result from methylation-associated mutational processes (e.g., C→T transitions at methylated CpGs) rather than being the primary drivers of methylation divergence. While additional analyses may not be necessary, explicitly acknowledging these alternative explanations would strengthen the interpretation.

Regarding the discussion comparing the distance between CpG-disrupting SNVs and trans-DMRs, without information on the absolute or relative distance distributions, it was difficult to assess the magnitude of the observed differences. Moreover, trans-DMRs, by definition, are not driven by local (cis) variation, and the lack of proximity to CpG-disrupting SNVs is expected. Clarifying what additional insight this analysis provides beyond this expectation may improve this section.

One potential extension would be to examine whether the same cis-acting SNVs are consistently associated with methylation differences across multiple cell types. If these variants are mechanistically causal, one might expect their effects to be reproducible across contexts, or at least more frequent than expected by chance. Such an analysis could further support the proposed link between sequence variation and methylation divergence.

Regarding their two-step sign test analysis, because enrichment-based approaches can sometimes overemphasize statistical significance without reflecting effect size, I wonder if incorporating the magnitude of methylation change would provide additional information or strengthen these findings. While the authors highlight some cases, such as TUBB2 and GRIK, a more general overview and/or integration of effect size into the analysis or discussion would improve interpretability.

Reviewer #3 (Public review):

Summary:

Ma et al. use human-chimpanzee tetraploid cells to examine species differences in DNA methylation. They identify differentially methylated regions under cis or trans regulation. Cis-DMRs are enriched near SNVs that disrupt or create CpGs, providing a plausible mechanism for cis changes in methylation. They also seek to identify transcription factors that might affect methylation in trans, as well as gene sets with evidence for consistent changes in methylation and expression between humans and chimpanzees.

Strengths:

The authors have generated a new dataset across multiple cell types, examining differences in DNA methylation between humans and chimpanzees using human diploid cells, chimpanzee diploid cells, and human-chimpanzee tetraploid cells. Using this dataset, they identify that cis-DMRs are enriched near SNVs that disrupt or create CpGs compared to trans-DMRs, and identify transcription factors as candidate trans-acting factors. Both identified SNVs and transcription factors are good candidates for future experimentation. Further, they find that cis-DMRs are more highly correlated with cis-expressed genes than trans-DMRs with trans-expressed genes, providing evidence that methylation and expression are linked genome-wide.

Weaknesses:

The authors could greatly improve the manuscript by focusing on two issues.

(1) Strengthening their cis/trans analysis, including:

a) only showing or analyzing genomic regions that pass FDR correction;

b) clarifying how cis genes are defined (Figure 2B shows some genes labeled as cis where the direction-of-effect differs between hybrid and parent cells);

c) assessing how well powered they are to perform each analysis.

(2) Softening claims about human evolution or human specificity for several reasons:

a) Their comparison lacks tetraploid controls (e.g. human-human tetraploids and chimp-chimp tetraploids) or experimental follow-up in diploid cells, making it hard to be certain that observed effects are not due to ploidy.

b) There are no outgroup species included in the analysis.

c) The use of no or very loose FDR corrections with the sign test makes it difficult to draw conclusions.

d) Experimental data to link SNVs to changes in cis methylation or identified transcription factors to changes in trans methylation would be needed to validate the authors' predictions.

Author response:

Summary of changes:

(1) Framing of the sign test results. Throughout the manuscript we have clarified that the sign test provides direct evidence that the implicated gene sets are under lineage-specific selection, while the connection of this selective signal to any specific human phenotype is inferential and would require experimental validation. We also now emphasize that some sign test results are expected to be false positives, as indicated by the FDR.

(2) FDR and cis:trans quantification. We have added explicit statements that (i) the cis dominance is consistent across a range of FDR thresholds (FDR < 0.25, 0.20, 0.10, 0.05, 0.01), (ii) the mean proportion of divergent sites explained by cis-regulation is 91%, and (iii) using the changepoint detection method, we identified 9159 cis-DMRs and 2046 trans-DMRs in total for downstream TF motif and SNV analysis. A power analysis comparing sensitivity to detect cis vs. trans effects is provided as author response image 1.

(3) Clarification of regulation groups, including pure cis, pure trans, conserved, cis × trans and cis + trans. We have added explicit definitions of these categories in the Results.

(4) Cis-gene direction-of-effect definition. We have clarified in the text that cis-regulated genes are defined as those where the direction of allelic difference is consistent between hybrid and parental samples. We note that the scatter plots in Figure 2B as well as Figure 2 supplemental figures plots the per-sample average percent difference in methylation. Cis genes appearing to have opposite direction-of-effect along the y=x diagonal on the scatter plot were due to a difference between this average, used only for visualization, and the values used for analysis in our statistical model. For values close to zero, average log2FC or per cent difference can flip sign even when the model correctly identifies the site as cis-regulated. The regulation classification is based on a statistical model and not the direction of average difference, so there will naturally be some small deviation around the axes when visualizing the results as a scatter plot.

(5) New limitations. We have added four new limitations to the Discussion: (i) absence of an outgroup species; (ii) TFBS-disrupting mutations as an alternative cis mechanism; (iii) the need for experimental validation of SNV-methylation and TF-methylation links; and (iv) the scope of cell types examined.

(6) Quantitative reporting. We have replaced "substantial" with specific numbers throughout.

(7) Additional supplemental resource. A list of cis–cis regulated promoters and their associated genes has been added as Supplemental File 8.

(8) Minor corrections. Missing citation for Hallgrímsdóttir et al. (2024) added; "Preprint at" removed from published paper citations; sequencing coverage reported in Methods.

eLife Assessment:

This study presents an important examination of the role of cis-acting versus trans-acting genetic variation on DNA methylation divergence between humans and chimpanzees, including its consequences for gene expression. By differentiating fused interspecies tetraploid cell lines into multiple cell types, the study provides compelling evidence for the importance of cis-acting changes, but incomplete evidence that these changes are of importance for adaptive trait evolution in humans. This work will be of interest to biologists and evolutionary anthropologists studying the evolution and genetics of gene regulation, particularly in primates.

We appreciate the assessment and agree that the evidence for adaptive trait evolution is indirect. We would like to clarify what the sign test does and doesn’t establish. By rejecting a rigorous neutral null model, the sign test provides strong evidence for lineage-specific selection on DNA methylation divergence. Since the coordinated directional divergence we observe is not expected under neutral evolution, we are confident that the data are evidence of selection. However, the sign test does not identify the specific traits that selection acted upon; the connection between the signal of selection and any phenotype is speculative. Accordingly, we have now revised the manuscript to describe the sign test results as evidence of selection while keeping the interpretation of specific traits explicitly speculative. We also added discussion of the limitations on the trait-level inference (such as the absence of an outgroup, reliance on clinical HPO ontologies, etc.).

Reviewer #1 (Public review):

Ma et al. use human-chimpanzee tetraploid cells across different cell types to identify the genetic causes and then transcriptomic consequences of divergence in DNA methylation. They conclude that the evolution of DNA methylation is driven primarily by cis-regulatory changes, and that the evolution of CpG sites contributes to cis-regulation, while transcription factor expression underlies some trans changes. They then argue that divergence in DNA methylation is associated with changes in gene expression and may contribute to human phenotypes.

We thank Reviewer 1 for a detailed summary of our work.

The tetraploid model is able to provide compelling evidence that most regulatory evolution occurs due to cis-regulatory changes. My only concern is that the extent of trans-changes may be overstated, as almost all are eliminated by changing from a nominal p-value criterion to even a 25% false discovery rate.

We agree that trans-effects are more sensitive to FDR thresholds than cis-effects, and we have addressed this directly. This is a well-recognised limitation of existing cis/trans classification frameworks, and we acknowledge this limitation in the discussion and have added suggestions in experimental set-up to mitigate it. To address the reviewer's concern directly, we have added a formal power analysis quantifying the relative power to detect cis and trans effects as a function of read depth as well as biological noise (dispersion). We now explicitly state in the Results that the cis dominance over trans is consistent across a range of FDR cutoffs (FDR < 0.25, 0.20, 0.10, 0.05, 0.01; Figure 2—figure supplement 1–2), and we note that the greater FDR-sensitivity of trans-DMRs likely reflects systematically smaller effect sizes rather than false positives. We have added a power analysis in Author response image 1 that formally compares our sensitivity to detect cis vs. trans effects. We also note that DMR-level analyses recover both cis and trans effects at meaningful FDRs, consistent with the idea that trans effects are real but requires more replicates as well as a higher sequencing depth to make significant calls at the same level as cis changes at the individual CpG level.

Power analysis of cis and trans effects detected by the beta-binomial GLM framework.

For each combination of effect size, per-unit read depth, and overdispersion (φ), we simulated a genome of 10,000 units — either individual CpG sites or regions/DMRs — comprising 2,000 true-effect units and 8,000 null units. The simulation is unit-agnostic: a "unit" represents a single CpG when modeling the per-CpG analysis and an aggregated region when modeling the regional (DMR-level) analysis, with read depth interpreted as the per-CpG coverage or the region-level (aggregate) coverage, respectively, and effect size as the corresponding per-CpG or region-level H–C methylation difference.

Effects were classified using the same two-model scheme as the main analysis. A unit was called cis if the hybrid allelic (H–C) test was significant and the allele-by-generation interaction was not significant, and trans if the interaction was significant and the hybrid allelic test was not significant; all tests were Benjamini–Hochberg–corrected across the genome at the stated FDR threshold. Each unit was simulated with the full six-measurement design (two parental samples and four hybrid alleles from two hybrids), both models were fit, each set of p-values was BH-corrected across the genome, and "power" is the fraction of true units that land in the correct class.

Cis-effect units were simulated with an equal H–C difference in both parents and hybrid (zero true interaction); trans-effect units with a parental H–C difference and no hybrid allelic difference (full-magnitude interaction); null units with no difference in either parents or hybrid. Sequencing depth was treated as a fixed design parameter: every sample at every simulated unit was assigned the same total read depth (the value on the y-axis of the heatmaps), corresponding to per-CpG coverage for the site-level analysis or aggregate per-region coverage for the regional analysis.

We estimated the beta-binomial overdispersion (φ) directly from the data for each cell type used in this study: DA = 0.024, CNCC = 0.016, IPSC = 0.013, SKM = 0.014, and HEP = 0.011 (median = 0.014). To span this empirical range and to assess robustness to higher-than-observed dispersion, we evaluated power at φ = 0.015 (approximating the empirically observed median), 0.05, and 0.10; the latter two represent substantially more conservative (higher-noise) scenarios than any cell type in our data. Because power declines monotonically with φ, the φ = 0.015 row reflects the sensitivity expected under realistic conditions, while the higher rows bound the worst case. For each unit, a methylation probability was drawn from a beta distribution whose mean equals the unit's true methylation level and whose dispersion is set by the overdispersion parameter φ, and the number of methylated reads was then drawn from a binomial with that probability and the specified total depth. Consequently, the reported power describes the sensitivity expected at a given uniform per-unit (per-CpG or per-region) depth, spanning 10× to 200×, and is interpreted as the proportion of true-effect units assigned to the correct class.

The follow-up analyses are incomplete with major gaps. The authors focus on single potential mechanisms for cis- and trans-changes, but it is not clear to what degree these mechanisms explain the extent of cis and trans changes. There are also other mechanisms which are not investigated, such as the importance of TF binding sites for cis-regulatory evolution.

We agree that CpG-disrupting SNVs and TF binding motifs are just two out of many possible mechanisms, and we have added explicit acknowledgements of this in the text. Our goal was not to investigate all possible mechanisms, but rather to focus on two that showed promising results. For cis-regulation, we now state that TFBS mutations and other sequence variants could also contribute. We agree that further mechanistic investigation is an important direction for future work, and we have noted this explicitly in the Discussion.

Next, the authors seek to show that differences in DNA methylation are functionally relevant... I worry that this result could be confounded by larger effect sizes for cis-changes than trans effects.

We agree that larger cis-regulatory effect sizes could contribute to this result, and have now mentioned this in the Discussion: “This finding suggests that the repressive effects of species-specific methylation are strongest when both layers of regulation are driven in cis, perhaps because cis-regulated methylation changes tend to have larger effect sizes than trans-regulated changes.”

Finally, the authors make a case for lineage-specific selection on DNA methylation that is connected to human traits. This evidence was not convincing. In fact, it is even said that these tests cannot be interpreted as evidence of lineage-specific selection (lines 399-401), so I am confused why these results are framed as testing for selection.

We appreciate this comment but want to clarify. Lines 399–401 in the original manuscript state that the dental pulp stem cell (DPSC) sign test results specifically should not be interpreted as evidence of lineage-specific selection, because DPSCs use parental rather than hybrid cell lines. This caveat applies only to DPSCs. For all other cell types (iPSC, DA, CNCC, SKM, HEP), which use hybrid cell lines that control for trans-acting confounders, the sign test does indeed provide evidence of non-random directional bias inconsistent with neutral evolution, thereby constituting evidence of lineage-specific selection. We have added a clarifying sentence to make this distinction explicit: "For all other cell types (iPSC, DA, SKM, HEP, CNCC), which use hybrid cell lines, the sign test results provide direct evidence of lineage-specific selection on the implicated gene sets, as the hybrid system controls for trans-acting confounders."

To be clear about the scope of the claim: the sign test directly demonstrates that the implicated gene sets are under lineage-specific selection. What remains inferential is the connection between this selective signal and any specific human phenotype, which would require experiments or further validation to establish; we have revised the framing throughout to keep this distinction explicit.

Reviewer #1 (Recommendations for the authors):

(1) Line 96: How many iPSC lines were used for each species? When a replicate is mentioned (e.g., Figure 1B), are those technical or biological replicates?

We have added the iPSC line counts to both the Results and Methods sections: two human and two chimpanzee iPSC lines were used. Replicates shown in Figure 1B are biological replicates (independent differentiation experiments from the same iPSC lines).

(2) Line 98: How many donors for the DPSCs?

We have added DPSC donor counts to the Methods: two human and two chimpanzee donors.

(3) Line 103: PCA and other analyses combined data from WGBS and RRBS cell types. Some clarification about how this was done (e.g. restricting to RRBS sites?) would be useful in the main text as well as an acknowledgement of potential biases. And

(4) Line 103: Related, what is the number of CpG sites considered?

We have added a clarifying sentence to the Results: cross-cell-type analyses including Figure 1B heatmap and Figure 1C PCA analysis were restricted to CpG sites covered in all samples (i.e., the intersection of all covered sites, after phasing the hybrid samples), totaling 561,762 shared CpG sites. Within-cell-type PCA analysis in Figure 1—supplemental figure 1 was restricted to CpG sites shared across samples of the same cell type, which contains 31,802,196 sites for CNCC samples, 26,163,237 sites for DA, 29,787,850 sites for DPSC, and 1,006,674, 935,636 and 952,835 sites for IPSC, SKM and HEP, respectively. We acknowledge that this restriction to the intersected CpG sites result in PCA performed on sites covered by both WGBS and RRBS, which are most frequent in CpG-dense regions.

(5) Line 103: For Figure 1B, it would be useful to first mention how human/chimp reads were parsed and for how many CpG sites this was possible.

We have added a brief description of the allele-parsing approach (SNPsplit using species-specific SNV positions). For hybrid WGBS samples (DA and CNCC), on average, 48.2% ± 0.2% of all aligned reads are unassignable using SNPs between human and chimpanzee genome, 25.9% ± 0.2% of the reads are specific to human, and 25.1% ± 0.2% of reads are specific to chimpanzee. For hybrid RRBS samples (IPSC, SKM and HEP), on average, 88.4% ± 0.3% of all aligned reads are unassignable using SNPs between human and chimpanzee genome, 5.3%±0.1% of the reads are specific to human; 5.3%±0.1% of reads are specific to chimpanzee. We have added information above to the main text. Per sample splitting reports containing the exact number of reads assignable to each of human and chimpanzee are provided as a comprehensive table at Supplemental File 1. The lower assignment rate in RRBS is expected, and we have performed two additional analyses to demonstrate why. First, the reduced representation strategy concentrates coverage in CpG-dense promoters that are highly conserved between human and chimpanzee and therefore SNV-poor, while the short MspI fragments that dominate these libraries frequently fail to reach an informative SNV even when one is nearby. Quantifying this directly: among the 16,845,376 CpG positions covered by the union of all unphased hybrid RRBS libraries, 65.1% contain a human–chimpanzee SNV within ±50 bp (the reach of a typical RRBS read), compared with 90.8% within ±150 bp (the reach of a typical WGBS read pair) — a 1.40-fold difference in SNV accessibility at the very same CpGs. Second, we compared the aligned-length distributions of allele-resolved and unassigned reads across all ten hybrid libraries (Author response image 2). In the four WGBS libraries the two distributions are indistinguishable, confirming that read length is not driving mappability. In contrast, for all six RRBS libraries, allele-resolved reads are consistently longer than unassigned reads, with the effect reproducible across every cell type and replicate.

Aligned read length determines allele-assignment success in RRBS but not in WGBS.

Aligned read lengths were extracted from the Bismark-aligned BAM files produced by phasing all ten hybrid bisulfite libraries: four WGBS libraries (CNCC and DA, two replicates each; left) and six RRBS libraries (HEP, iPSC and SKM, two replicates each; right). For each library, reads assigned to either the human (genome1) or chimpanzee (genome2) alleles were pooled into a single phased set (blue) and compared with the reads that could not be phased (grey); distributions were computed from subsampled reads and are shown as boxplots (box, interquartile range; line, median; whiskers, 5th–95th percentiles). In the WGBS libraries, phased and unassigned reads have effectively identical length distributions (median 150 bp for both; means 147.4–147.8 vs 142.1–143.1 bp), indicating that read length does not drive mappability in WGBS. In the RRBS libraries, which are dominated by short MspI fragments, phased reads are consistently longer than unassigned reads across every cell type and replicate (medians 70–75 vs 49–50 bp; means 80.5–85.7 vs 58.6–60.6 bp, ~1.4-fold). Because a read can be assigned to a parental allele only if it spans an informative human–chimpanzee SNV, these results show that the short fragment length characteristic of RRBS — together with its enrichment for SNV-poor, evolutionarily conserved CpG-dense promoters — accounts for its lower allele-resolution rate relative to WGBS.

(6) Line 103: Hybrid/parental line isn't noted in the PCA. Is this just the parental samples?

We clarify that Figure 1C shows both parental and hybrid samples, and we have updated the figure caption for Figure 1C to note that both hybrid and parental samples are plotted in the PCA. The parent/hybrid stratifications are shown in Figure 1—figure supplement 1 and referred to in the main text.

(7) Line 149: No citation # is given for the Hallgrimsdottir paper.

Corrected. Citation number 50 (Hallgrímsdóttir et al., 2024) has been added.

(8) Line 167-169: Some explicit test here would be useful. Is this more than we'd expect just because the majority of CpG sites have conserved methylation levels? How much more do divergent CpG sites cluster than expected by chance?

We agree that an explicit test of clustering would be informative, and we want to be clear about what our analysis does and does not establish. The changepoint algorithm is not a test of non-random clustering: it detects shifts in the local composition of regulatory classifications along the genome and segments each chromosome into locally homogeneous stretches. Its purpose here is enrichment rather than inference, because it selects regions in which neighboring CpG sites share both a regulatory classification and a consistent direction of species bias, so that the downstream motif-enrichment and CpG-SNV proximity analyses operate on interpretable units rather than on mixtures of mechanisms. Because the segmentation is not compared against a null model, it cannot by itself establish that divergent sites co-localize more than expected given that 83–93% of all sites are conserved. We have therefore revised the text so that Figure 2D reads as a descriptive summary of the neighbourhood composition of each regulation group, and removed any implication that the co-localization has been tested against a chance expectation. Asking whether divergent sites cluster more than expected is a genuinely different question from the one our DMR-calling addresses and answering it well would require a dedicated null model. We think this is a worthwhile analysis and a natural extension of this framework, but it is separate from the claims we make here, and we have not attempted it in this revision.

(9) Line 186-187: It should be made clear that you looked to cis, trans, and cis + trans DMRs independently.

We have revised this sentence to make explicit that cis-DMRs, trans-DMRs, and cis+trans DMRs were identified independently using the changepoint detection method applied separately to each regulatory class.

(10) Line 190-192: Some quantification of the degree to which cis effects exceed trans effects would be useful. What proportion of total divergence (e.g. mean and s.d.) is explained by cis factors?

We have added this quantification: “At FDR < 0.05, cis-regulation accounted for a mean of 91% of divergent sites and trans-regulation a mean of 4.5% across cell types, and cis remained the dominant category at every FDR cutoff tested (FDR < 0.25–0.01; Figure 2C, Figure 2—figure supplement 1–2, Supplemental Files 1–2).”

(11) Line 194-196: It should be made clear that motif enrichment was done in the human genome, right? Are the results consistent if motif enrichment was done in the chimpanzee genome?

We have clarified in the Methods that HOMER motif enrichment was performed using the human genome (hg38) as background. Repeating the analysis using the chimpanzee genome as background is a valuable future direction; we acknowledge this has not been formally tested.

(12) Line 208: For TFs other than FOXM1 and FOXA2, is TF expression generally associated (positively or negatively) with trans DMRs favoring the chimpanzee or human lineage? An idea of the distribution would be useful.

We plotted the distribution of differential TF expression for those with predicted binding sites enriched in Hu>Ch trans-DMRs versus Hu<Ch trans-DMRs in DA and CNCC. In neither cell type do the two distributions show a clear separation of directional shift. TF expression differences are broadly overlapping regardless of whether the associated trans-DMR favors the human or chimpanzee lineage. We speculate the lack of separation reflects multiple trans-acting inputs shaping methylation divergence such that single TF expression does not predict it. Aggregating across TFs with potentially opposing effects could further contribute to this observation as well (Figure 3—figure supplement 1). FOXM1 and FOXA2 are individual cases where a suggestive relationship was visible rather than evidence of genome-wide pattern.

Distribution of differential expression of TFs with predicted binding sites enriched in trans-DMRs.

Density distribution of TF expression difference in human versus chimpanzee parental samples. There is a lack of separation between expression of TFs with predicted binding sites enriched in Hu>Ch trans-DMRs versus Hu<Ch trans-DMRs, in both DA and CNCC.

(13) Line 211-212: Stats are needed.

This statement is intended as a qualitative description rather than a quantitative statement. With only 4 cell types per comparison (Figure 3B), we do not have enough statistical power to establish a significant correlation, therefore we present this result as an exploratory observation rather than a statistically significant finding. We have therefore removed "consistent with this expectation," reframed the FOXM1 and FOXA2 results as descriptive observations from individual factors, and added an explicit statement that they are exploratory and hypothesis-generating rather than evidence of a genome-wide pattern.

(14) Figure 3C: Direction is not given. It should be made clear that this is a gain/loss in humans.

We have updated the Figure 3C caption to explicitly label the three SNV categories as: CpG gains in humans (human allele creates a CpG), CpG losses in humans (human allele disrupts a CpG), and no CpG change.

(15) Figure 3C: Most cell types have a bias towards increased DNA methylation when there are no nearby CpG sites (gray bars). Why? Is this a consequence of a reference bias?

We thank the reviewer for raising this. The grey bars correspond to methylation levels of conserved CpG sites near SNVs with no CpG-altering substitution and serve as the genome-wide background/control category, against which the substitution-containing categories are compared. Their modest skew toward higher methylation reflects a global offset shared by all categories rather than a substitution-specific effect. We suggest two non-exclusive explanations for this baseline offset: a genuine genome-wide difference in the distribution of human- versus chimpanzee-biased methylation, and/or a possible contribution of reference genome mapping bias arising from alignment to the human genome. Importantly, this does not affect the conclusions drawn from Figure 3C. Because the genome-wide offset is shared across all categories, the substitution-attributable signal is reflected in the difference between the CpG-gain/loss categories and the no-change (grey) baseline, and any genome-wide baseline shift — biological or technical — cancels in that comparison. We therefore interpret the grey-category asymmetry as a background property and explicitly do not attribute it to the substitutions themselves. Future work such as analyzing with reads mapped to the chimpanzee or a masked consensus genome or restricting to mappability- and coverage-matched regions could directly quantify any reference-bias contribution, and we note these as promising directions.

(16) Line 258-269: Figure 3D is not referenced in the text.

We have added a reference to Figure 3D in the relevant paragraph of the Results: “To test this, we compared the genomic distance between each DMR and its nearest CpG SNV. Consistent with our model, we found consistently shorter distances for cis-DMRs than trans-DMRs across all cell types examined: Median distances to the nearest CpG SNV were 24 ± 6 bp for trans-DMRs vs. 11 ± 2 bp for cis-DMRs (Figure 3D; Mann-Whitney p < 0.001 for all comparisons) [65].”

(17) It is unclear why Figure 4A-C focus exclusively on cis-regulated promoters. Is this for biological or technical reasons?

Both. Biologically, cis-regulated promoters are the most interpretable for assessing cell type-specificity, because cis effects are allele-intrinsic and not confounded by differences in trans-acting environments across cell types. Technically, trans-regulated promoters show methylation differences that depend on the trans-environment, making cross-cell-type comparisons less straightforward.

(18) Line 371: How many DMRs are being considered?

We have added the total number of DMRs (promoters) considered in the sign test analysis: 6,283.

(19) Line 371-372: What is the ASM/ASE concordance test? I thought you were restricting the analysis to genes with a negative association between ASM and ASE (lines 366-367), so this needs to be clarified.

We have clarified this in the text. The two-step procedure works as follows: (1) a binomial sign test is applied to all genes with promoter ASM to identify gene sets with directional bias in methylation; (2) separately, a binomial sign test is applied to the subset of genes where ASM and ASE show a negative (repressive) relationship. Gene sets that pass both steps are reported as candidates. The "ASM/ASE concordance test" refers to step 2. We have revised the text to make this two-step logic clearer.

(20) Line 379: It may be worth mentioning that these are the DMRs which are most likely to be functionally relevant in shaping expression levels.

We have added this sentence: " This subset represents cases where methylation changes are most likely to have a simple and direct repressive effect on transcription, filtering out more complex regulatory scenarios where methylation and expression changes may be uncoupled or subject to competing regulatory influences."

(21) Line 387: The test here needs to be clearer. Does a significant result mean an enrichment of differences in that gene set, or a bias in the ratio of Hu/Ch differences?

We have clarified: a significant result from the binomial sign test indicates a statistically significant bias in the ratio of human-biased to chimpanzee-biased changes within a gene set, relative to the genome-wide background ratio. It does not test for enrichment of the number of differences, only for non-random directionality.

(22) Line 403: Why the focus on expression, when the rest of the paper is focused on the methylation patterns?

Gene expression is the most direct functional readout of promoter methylation changes. The two-step sign test uses expression data in the second step precisely because it allows us to identify cases where methylation divergence has a demonstrable downstream consequence on transcription, which is the canonical mechanism by which promoter methylation affects phenotype. We have added a brief justification of this rationale to the text.

(23) Line 456: Not clear where there is actually evidence of lineage-specific selection.

We have added an explicit sentence in the Discussion describing what the evidence consists of: "Specifically, the evidence rejecting a neutral null model of cis-regulatory evolution consists of statistically significant directional bias in both promoter methylation and gene expression within functionally coherent gene sets, assessed using a two-step binomial sign test with permutation-based FDR correction." This constitutes direct evidence that the implicated gene sets are under lineage-specific selection; the separate question of which specific human phenotypes that selection shaped remains inferential and is framed as such throughout.

(24) Line 457-459: I think this line needs to be further explained: "Our use of hybrid cell lines was critical to this discovery, as it allowed us to isolate cis-acting regulatory changes from confounding trans-acting and environmental effects." Specifically, it's not apparent why cis-regulation is important for demonstrating evidence of selection.

We have added an explanation: "Isolating cis-acting changes is critical because the sign test requires that each promoter represents an independent observation; in hybrid cells, each allele's methylation is determined by its own cis-regulatory sequence, making observations across promoters genuinely independent and thus satisfying the assumptions of the binomial test."

(25) Line 460-465: The concluding claims throughout this paragraph appear to be overstated.

We have revised the concluding paragraph of the Results and the Discussion conclusion to soften the language.

(26) Line 487: Provide the actual numbers instead of saying "substantial".

We have replaced "substantial numbers of allele-specifically methylated promoters" with the actual count: 6,283 allele-specifically methylated promoters across all cell types.

(27) Line 502: I question this interpretation. While it could be that there is a shared mechanistic basis, it could also be that the cis effects tend to be larger than the trans effects, providing additional power to identify strong correlations.

We agree and have added this alternative explanation explicitly to the Discussion (see response to public review comment above).

(28) Line 510: But the results section states that it's not possible to interpret these as lineage-specific selection.

The caveat at lines 399–401 applies specifically to DPSCs, not to hybrid cell types. We have revised the text to make this distinction unambiguous.

(29) Line 601: What coverage was the methylation data?

We have added sequencing coverage and the number of CpG sites assayed below “Bisulfite-seq library preparation, sequencing and mapping” section in Methods. WGBS libraries achieved a mean coverage of 9.68× per CpG site, covering 43.6M unique CpG positions in CNCC (hybrid) and 43.4M in CNCC (parental), 41.9M in DA (hybrid) and 38.8M in DA (parental). RRBS libraries achieved a mean coverage of 5.02× per CpG site, covering 5.0M, 4.8M, and 4.5M unique CpG positions in the hybrid HEP, IPSC, and SKM libraries respectively, and 3.3M, 3.4M, and 3.2M in the corresponding parental libraries. The total number of CpG sites (union) of all WGBS samples is 48,994,320 sites, and the union of all CpG sites of RRBS samples is 7,289,251. The union of all samples (including WGBS and RRBS) is 49,049,052. We have also included coverage information per sample as a table in Supplemental File 1.

(30) Line 820: citation shouldn't include "preprint at".

Corrected. "Preprint at" has been removed from all published paper citations in the reference list.

Reviewer #2 (Public review):

This manuscript investigates the causes and consequences of human-specific DNA methylation divergence relative to chimpanzees... This study provides a valuable dataset and a compelling framework for understanding how local sequence variation contributes to epigenetic and transcriptional divergence, with likely broad impact in comparative and evolutionary genomics.

We thank Reviewer 2 for this positive assessment and for the constructive suggestions.

Although the authors identify transcription factors associated with differential methylation, it is unclear what proportion of differentially methylated CpGs or DMRs can be attributed to these factors. Providing a quantitative estimate would help assess the relative contribution of trans-acting regulation.

We agree that a quantitative estimate of the proportion of differentially methylated CpGs or DMRs attributable to these transcription factors would be informative in principle. In practice, however, such an estimate is challenging to produce reliably: predicted binding sites are not all occupied or functional, individual DMRs are typically influenced by multiple trans-factors, and we lack the per-site functional data needed to assign methylation differences to specific TFs with confidence. More importantly, even if we could generate such a number, an association-based estimate would not establish that these factors causally drive the observed methylation differences. We have added a note to the main text clarifying that our analysis identifies candidate trans-acting factors associated with differential methylation rather than quantifying their causal contribution.

The analysis of CpG-disrupting mutations is interesting but raises two concerns. First, other classes of variants—such as transcription factor binding site-disrupting mutations—could also influence local methylation patterns and are not considered here. Second, the causal direction remains ambiguous: CpG-disrupting mutations may result from methylation-associated mutational processes (e.g., C->T transitions at methylated CpGs) rather than being the primary drivers of methylation divergence.

We have addressed both concerns in the revised manuscript. On the first point, our goal was not to suggest that CpG-disrupting variants are the only causes of local methylation divergence. We now explicitly acknowledge in the Results that TFBS mutations and other variant classes could also contribute to cis-methylation divergence and represent an important avenue for future investigation. On the second point: while CpG-disrupting mutations are indeed expected to be enriched in highly methylated regions, they would be expected to associate with methylation level rather than methylation divergence, and thus cannot explain the patterns we observe.

Regarding the discussion comparing the distance between CpG-disrupting SNVs and trans-DMRs, without information on the absolute or relative distance distributions, it was difficult to assess the magnitude of the observed differences. Moreover, trans-DMRs, by definition, are not driven by local (cis) variation, and the lack of proximity to CpG-disrupting SNVs is expected. Clarifying what additional insight this analysis provides beyond this expectation may improve this section.

We agree that the lack of proximity of trans-DMRs to CpG-disrupting SNVs is expected. We have revised the text to clarify that this analysis serves as an affirming result: it is consistent with the fact that cis-DMR are under local sequence-driven regulation, and that the CpG-disrupting SNV mechanism is more closely associated to cis-DMRs than trans-DMRs. The quantifications for distance distributions are shown in Figure 3D, and we have added a sentence noting the median distances for trans-DMRs vs. cis-DMRs: 24 ± 6 bp vs. 11 ± 2 bp.

One potential extension would be to examine whether the same cis-acting SNVs are consistently associated with methylation differences across multiple cell types.

We thank the reviewer for this suggestion and have performed this analysis below. For each of the three regulatory region classes (promoters, enhancers, CTCF binding sites), we intersected CpG SNVs (±100 bp window) with cis-classified DMRs (pure_cis or cis_plus_trans at BH-corrected FDR < 0.05) across the five cell types (CNCC, DA, HEP, IPSC, SKM), and counted the number of cell types in which each SNV overlapped a cis-DMR. To assess significance, we compared this distribution to a permutation null in which, for each cell type, we randomly sampled the same number of regions from that cell type's pool of tested-but-conserved regions of the same class (1,000 permutations); SNVs were restricted to those falling within the tested universe across cell types.

Local methylation effects of CpG SNVs were shared across cell types substantially more than expected by chance in all three region classes. The observed mean number of cell types sharing per SNV-in-cis-DMR was 1.244 vs. 1.077 ± 0.006 in the null for promoters, 1.133 vs. 1.042 ± 0.001 for enhancers, and 1.171 vs. 1.061 ± 0.002 for CTCF binding sites (all p < 0.001). Fold-enrichment scaled sharply with the degree of sharing: at the highest sharing level (cis-DMR in all five cell types), observed counts exceeded the permutation null by ~3,500-fold for promoters, ~800-fold for CTCF, and ~700-fold for enhancers. The slightly stronger consistency signal at promoters relative to enhancers is consistent with the more constitutive activity of promoter elements across cell types. These results indicate that the same cis-acting variants tend to drive methylation divergence across multiple cell types, supporting a shared mechanistic basis for cis-methylation divergence. We have added the figure in the main text as Figure 3—figure supplement 2.

Regarding their two-step sign test analysis, because enrichment-based approaches can sometimes overemphasize statistical significance without reflecting effect size, I wonder if incorporating the magnitude of methylation change would provide additional information.

We agree. The magnitude of methylation change (effect size) for each gene and gene set is now provided in Supplemental File 7 alongside the directional bias statistics. We have added a sentence to the Results pointing readers to this resource.

Strength:

I recommend that the authors provide a list of cis-cis-regulated promoters and their associated genes, which would be a valuable resource for the field.

We have added this list as Supplemental File 8.

Reviewer #3 (Public review):

Ma et al. use human-chimpanzee tetraploid cells to examine species differences in DNA methylation... Both identified SNVs and transcription factors are good candidates for future experimentation. Further, they find that cis-DMRs are more highly correlated with cis-expressed genes than trans-DMRs with trans-expressed genes, providing evidence that methylation and expression are linked genome-wide.

We thank Reviewer 3 for this summary and for the detailed recommendations.

Weakness

(1) Strengthening their cis/trans analysis, including: (a) only showing or analyzing genomic regions that pass FDR correction; (b) clarifying how cis genes are defined (Figure 2B shows some genes labeled as cis where the direction-of-effect differs between hybrid and parent cells); (c) assessing how well powered they are to perform each analysis.

(a) Figure 2B has been updated to show regulatory classifications of promoter regions in cranial neural crest cells (CNCCs) at FDR < 0.05. Scatter plots and bar plots for all cell types at the level of individual CpG sites have been moved to Figure 2—figure supplement 1, and the corresponding promoter-level plots for all cell types other than CNCCs to Figure 2—figure supplement 2. We have clarified in the Figure 2—figure supplement 1 caption that nominal p-values are used to preserve the full dynamic range of the data, whereas FDR-corrected thresholds are applied for visualization of promoter scatter plots as well as all formal inference and for reporting significant candidate loci. FDR-corrected results are presented in Figure 2—figure supplements 1–2, and we have added explicit statements in the Results confirming that the cis: trans ratio is stable across FDR thresholds.

(b) We have clarified the definition of cis-regulated genes in the text: cis-regulated genes are defined as those where the direction of allelic difference is consistent between hybrid and parental samples for both methylation and expression. We have verified that this definition is applied consistently throughout the manuscript and downstream analyses.

(c) We have clarified the FDR correction approach for the sign test in the Methods: FDRs are estimated by permutation (10,000 permutations), and we report results at FDR < 0.25 for the two-step test with supporting evidence from the methylation-only sign test (nominal p < 0.05). We acknowledge that this is a relatively liberal threshold. The sign test itself provides direct evidence that the implicated gene sets are under lineage-specific selection; what we have toned down is the downstream interpretation, framing the link between these gene sets and specific human phenotypes as inferential and requiring experimental validation rather than as confirmed.

(d) We agree that experimental validation would strengthen the mechanistic claims. We have added this explicitly to the Limitations section: "definitive demonstration of causality requires targeted experimental perturbations such as base editing (for CpG-disrupting SNVs) or TF knockdown/overexpression experiments (for trans-acting TFs)."

(2) Softening claims about human evolution or human specificity for several reasons: (a) Their comparison lacks tetraploid controls (e.g. human-human tetraploids and chimp-chimp tetraploids) or experimental follow-up in diploid cells, making it hard to be certain that observed effects are not due to ploidy. (b) There are no outgroup species included in the analysis. (c) The use of no or very loose FDR corrections with the sign test makes it difficult to draw conclusions. (d) Experimental data to link SNVs to changes in cis methylation or identified transcription factors to changes in trans methylation would be needed to validate the authors' predictions.

(a) The concern about ploidy effects is noted. We note, however, that the primary evidence for cis-regulation comes from allele-specific methylation within the same tetraploid nucleus, where both alleles experience identical ploidy and trans-environments. Ploidy effects would be expected to affect both alleles equally and would therefore not produce the allele-specific differences we observe. We acknowledge that single-species tetraploid controls would provide additional confidence and represent a valuable future experiment.

(b) We have added the absence of an outgroup species as an explicit limitation in the Discussion.

(c) We have clarified the FDR correction approach for the sign test in the Methods: FDRs are estimated by permutation (10,000 permutations), and we report results at FDR < 0.25 for the two-step test with supporting evidence from the methylation-only sign test (nominal p < 0.05). We acknowledge that this is a relatively liberal threshold and have emphasized this by adding a new sentence: “However, as indicated by the FDR, we do expect some fraction of these candidate gene sets to be false positives.”

(d) We agree that experimental validation would strengthen the mechanistic claims. We have added this explicitly to the Limitations section: "definitive demonstration of causality requires targeted experimental perturbations such as base editing (for CpG-disrupting SNVs) or TF knockdown/overexpression experiments (for trans-acting TFs)."

Reviewer #3 (Recommendations for the authors):

(1) Why are dental pulp stem cells so different from the other cell types transcriptionally (e.g. Figure 1C)? Is this due to being primary cells vs. stem-cell-derived cells?

The transcriptional distinctiveness of DPSCs is most likely attributable to their status as primary cells rather than iPSC-derived cells. Primary cells retain the epigenetic and transcriptional signatures of their tissue of origin and have not undergone reprogramming, whereas iPSC-derived cell types share a common pluripotent origin that may reduce inter-cell-type transcriptional distance. Additionally, DPSCs are mesenchymal stem cells with a distinct developmental lineage and were obtained from adult donors rather than differentiated in vitro, which may further contribute to their distinct profile. We have added a brief note to the Results acknowledging this.

(2) I am concerned about how cis genes are defined. In Figure 2B, there are genes labeled as cis, where the direction-of-effect differs between hybrid and parent cells. I believe cis should only include genes with the same direction-of-effect between hybrid and parent cells, as illustrated in Figure 2A. This fix should propagate through the ensuing figures and results.

We have clarified the definition of cis-regulated genes in the text (see response to Weakness 1b above). The classification framework assigns a site to "cis" when the allele effect test is significant in hybrids and the System × Allele interaction term is not significant, meaning there is no detectable change in the allelic difference between the hybrid and parental systems. Sites where the allelic difference genuinely differs between systems — that is, where the interaction term is significant — are classified as cis × trans (interactive) or cis + trans (additive) rather than pure cis, depending on whether the two systems disagree in direction or only in magnitude.

We note that Figure 2B plots parental and hybrid methylation differences as the per-sample average difference in fractional methylation, which is used for visualization only. Points labeled cis that fall on the opposite side of the y = x diagonal are an artifact of this averaging: where the underlying difference is close to zero, the coverage-weighted per-sample average can flip sign even though the beta-binomial model, which fits the read counts directly and accounts for coverage and overdispersion, finds no significant interaction. The regulatory classification is based on the model, not on the sign of the averaged difference, so a small amount of scatter around the diagonal is expected. We have verified that this definition is applied consistently throughout the manuscript and downstream analyses and have clarified the caption of Figure 2B to make this explicit.

(3) Are you equally powered to detect cis and trans effects?

At the single-CpG level, the two hypotheses do differ in statistical power, but at the regional level this difference is mitigated by the higher effective read depth, and we now state this explicitly in the revised manuscript.

The first hypothesis in our classification framework tests allele-specific methylation directly, whereas the second hypothesis includes an interaction term that evaluates whether allelic differences shift between the parental and hybrid systems. Because the interaction test is inherently less powerful, it requires greater sequencing depth and, ideally, additional replicates to yield comparably confident estimates and to assign sites to certain classification categories with significance. As a consequence, at the single-CpG level, trans effects are more sensitive to FDR thresholds than cis effects of equivalent nominal significance, which reduces our power to detect them. At the regional level, however, as well as in CpG-level analyses with high-coverage samples, this disparity in power between cis and trans largely disappears.

This is the technical component of the issue. From a biological standpoint, cis effects are also intrinsically easier to detect than trans effects: estimates of trans effects necessarily incorporate not only trans-acting molecular mechanisms but also environmental and technical noise, so cis estimates will always be cleaner. This is a well-recognized limitation of existing cis/trans classification frameworks, and we acknowledge this limitation in the discussion and have added suggestions in experimental set-up to mitigate it. To address the reviewer's concern directly, we have performed a formal power analysis quantifying the relative power to detect cis and trans effects as a function of read depth (See Author response image 1).

(4) Are the results shown for SNVs in CpGs in Figure 3C only for cis-DMRs or for all DMRs? If the latter, please show cis-DMRs and trans-DMRs separately and explain why trans-DMRs show the same pattern as cis-DMRs. In addition, I would expect that the CpG that contains an SNV leading to CpG loss or gain would be the most affected in terms of methylation status, rather than CpGs within 25bp vs. 50bp vs. 100bp. It would be helpful to see that plot as well.

We thank the reviewer for the opportunity to clarify. Figure 3C is not an analysis stratified by cis-DMRs or trans-DMRs. Rather, it examines local methylation changes at conserved CpG sites located within 25 bp, 50 bp, and 100 bp of a CpG-disrupting SNV. The purpose of this analysis is to test a specific mechanistic hypothesis: that a CpG-disrupting mutation may influence the methylation status of nearby, intact CpG sites in cis. The analysis is therefore agnostic to cis/trans-DMR classification, as it asks a different question, namely whether sequence disruption at one CpG propagates locally to neighboring CpGs.

We note that Figure 3D addresses a related but distinct question, showing that cis-acting sequence changes more broadly (not restricted to CpG-disrupting mutations) are more strongly associated with cis-DMRs than with trans-DMRs. Together, Figures 3C and 3D support complementary aspects of the cis mechanism: the local propagation of methylation changes around disrupted CpGs (3C) and the broader association of sequence variation with cis-DMRs (3D).

Regarding the reviewer's second point: we agree that the CpG site directly disrupted by the SNV would itself be the most affected, by definition, since the CpG dinucleotide is destroyed (in the case of CpG loss) or newly created (in the case of CpG gain). Because methylation cannot be measured at a CpG that no longer exists (or has only just been created on one allele), this site cannot be included in a comparative methylation analysis across alleles or species. The 25/50/100 bp windows were therefore chosen specifically to test the local effect on conserved, measurable CpGs surrounding the disruption. We have clarified this rationale in the revised text/figure legend.

(5) The finding that cis-DMRs and cis-expressed genes are more correlated than trans-DMRs with trans-expressed genes is nice to see. What is a potential explanation for why the patterns across the genomic interval surrounding the TSS are so different across cell types, and even positively correlated near the TSS in dopaminergic neurons? Why do you not see a difference between cis and trans in CNCCs? In addition, it would be nice to see the correlation between cis-DMRs and trans-expressed genes, and vice versa, as a control for these plots.

We have added the suggested control analysis. For the cis-DMR/trans-expressed-gene and trans-DMR/cis-expressed-gene sets in both DA neurons and CNCCs, the correlations around the TSS are less consistently negative than for the cis-DMR/cis-expressed-gene set. In both cell types, cis-methylation and cis-expressed genes show the strongest opposing (negative) relationship to one another, which supports the specificity of the cis–cis coupling reported in Figure 4G. Regarding why the correlation profiles across the TSS-flanking interval differ so markedly across cell types—including the positive correlation near the TSS in dopaminergic neurons—and why the cis–trans distinction is less pronounced in CNCCs, we are not able to offer a confident explanation. The variation may reflect cell-type-specific regulatory architecture, but a great deal remains unknown about how methylation and transcription are coupled in a cell-type-specific manner, and our four cell types do not allow us to distinguish among the possible explanations. We think this is an open and interesting question, and that analyzing additional cell types with matched WGBS and gene expression data would be the most informative way to establish how general these patterns are and what drives the between-cell-type variation. We have noted this explicitly in the main text.

Spearman correlation of expression and methylation for CNCC and DA for subsets of genes with cis-regulated methylation and trans-regulated expression (orange), as well as genes with trans-regulated methylation and cis-regulated expression (blue).

Two kb surrounding each transcription start site (TSS) is shown, split into 10 bins of 200 bp where individual CpG sites are pooled and averaged as fractional methylation values and correlated with expression values of the neighboring genes (in TPM).

(6) The use of the sign test to identify pathways with the same direction of effect is a great approach. However, I am concerned with the use of no or very loose FDR corrections (e.g. FDR < 0.25 in line 372 and no FDR correction in line 374). Combined with Figure 6, where many of the signals appear to be weak or driven by a few genes (e.g. highly arched eyebrow and poor speech), it makes it difficult to draw robust conclusions. It is notable that the strongest signal is seen in DPSCs, which the authors note contains experimental confounds. We also think that the authors should tone down their biological interpretations, particularly for stem cells. For instance, COLEC11 is involved in neural crest cell migration but is identified from the iPSC analysis and not the CNCC analysis. Why not?

We have added a sentence to emphasize the possibility of some false positive sign test results (“However, as indicated by the FDR, we do expect some fraction of these candidate gene sets to be false positives”), and we have toned down the biological interpretations accordingly.

We want to clarify that the sign test directly demonstrates that these gene sets are under lineage-specific selection. We have clarified that the test does not establish a definitive link between these gene sets and specific human phenotypes; that link is inferential and would require experiments or further validation, and we have toned down the biological interpretations accordingly.

Regarding COLEC11: in response to this comment, we directly examined COLEC11 expression and methylation in CNCCs and found that it is in fact significant in this cell type. In CNCC hybrid cells, COLEC11 shows significant chimpanzee-biased allele-specific expression (log2 fold change [human/chimpanzee allele] = –2.45; DESeq adjusted p = 0.005), together with significant human-biased methylation (fractional methylation [human – chimpanzee] = 14%; adjusted p < 0.001). The direction of this divergence is therefore consistent with IPSC. The reason COLEC11 did not surface at the pathway level in CNCCs is that other genes in the same pathway contribute to the directional signal for that cell type; we would not expect every individual gene to emerge in the cell type where its pathway is most relevant. We have added a note making this point and cautioning against over-interpreting any single gene example.

(7) We appreciate that the authors provide a careful list of limitations of their study. In addition to the limitations listed, the authors should also include that experiments would be needed to validate that SNVs disrupting or creating CpGs affect methylation.

We have added this to the Limitations section: “definitive demonstration of causality requires targeted experimental perturbations such as base editing (for CpG-disrupting SNVs) or TF knockdown/overexpression experiments (for trans-acting TFs).”

(8) We also caution the authors in describing their findings as "human-specific" because there is no outgroup in this study.

We agree and have added the absence of an outgroup as an explicit limitation in the Discussion.

We thank all three reviewers again for their thorough and constructive engagement with our work. We believe the revised manuscript is improved as suggested.

Version 2

Abstract

The vast collection of human-specific traits– such as our unique morphology, cognition, behavior, and diseases– has long been attributed to gene expression divergence between us and our closest living relatives, chimpanzees. Theory suggests that changes to cis-regulatory elements such as promoters and enhancers may drive evolutionary adaptation, and DNA methylation is a key factor in transcriptional cis-regulation. However, we still lack an understanding of 1) how species-specific methylation patterns arise; 2) their downstream effects; and 3) whether they are a common target of natural selection. In this study, we investigated these three questions. By combining a novel hypothesis testing framework with DNA methylation data from six human and chimpanzee cell types, as well as fused interspecies hybrid cells, we disentangled cis– vs. trans-acting methylation divergence across the genome. Across cell types, we found that methylation divergence is primarily driven in cis, which can be linked in some cases to nearby sequence variants such as CpG gains and losses. Although less common, regions with trans-acting methylation divergence were enriched for specific transcription factor (TF) binding motifs, suggesting a role of TFs such as FOXM1 in these differences. Having established these causes of methylation divergence, we then examined the functional consequences of differential methylation. Although methylation lacks a consistent relationship with transcription, we observed that associations between methylation and gene expression are stronger for genes with cis-regulatory divergence. Moreover, we identified lineage-specific selection shaping promoter methylation at the level of entire pathways including those associated with human-specific traits such as speech, cognition, and susceptibility to infection with hepatitis C. Collectively, our findings provide a mechanistic framework suggesting that DNA methylation may occupy a key position, mediating the effects of both cis– and trans-acting factors on transcriptional networks, including those contributing to human-specific traits.

Introduction

While gene expression divergence has long been considered the primary driver of human evolution, identifying the molecular mechanisms underlying uniquely human traits remains challenging, particularly when they involve epigenetic modifications that control cell type-specific expression patterns1–4. Theory suggests that cell type-specific cis-regulatory variants may fuel evolutionary adaptation through their precise effects on individual genes within specific cell types, avoiding potentially deleterious consequences of broader trans-acting changes that affect multiple targets simultaneously5,6. However, while recent advances have enabled quantification of human-specific divergence in gene expression and chromatin accessibility, the role of DNA methylation—a key regulator of gene expression— in coordinating these changes and driving human-specific traits has not been fully explored6–10. DNA methylation, occurring primarily at CpG dinucleotides, is generally associated with transcriptional repression and provides stable, long-term regulatory control that has been linked to species-specific phenotypes in mammals11. Comparative methylome studies between humans and chimpanzees have revealed widespread tissue-specific methylation divergence, and linked human-specific hypomethylation to neurodevelopmental evolution and disease susceptibility7,12–22.

Studies of allele-specific expression (ASE) and chromatin accessibility measurements in interspecies hybrids provide a powerful approach for studying regulatory evolution by eliminating confounding factors inherent to cross-species comparisons, including differences in cell type composition, environmental conditions, and developmental timing23–32. Analogously, allele-specific methylation (ASM) in human-chimpanzee hybrid cells enables the isolation of cis-regulatory effects while controlling for shared trans-acting factors. Cis regulation of methylation is caused by factors such as sequence variants affecting methylation on the same allele as the variants, typically acting locally. In contrast, trans regulation of methylation involves diffusible factors such as DNA methyltransferases, transcription factors, or chromatin modifiers that can influence methylation patterns across many genomic loci (Figure 2A). Studying ASM enables direct quantification of the relative contribution of local versus diffusible factors to species-specific methylation patterns.

While previous studies of primate methylation evolution in post-mortem tissues and blood have provided valuable foundational insights7,12–16,18–22,33,34, several important questions remain to be addressed. These include: (1) generating comprehensive genome-wide methylation maps across multiple species and cell types; (2) distinguishing between cis-versus trans-regulatory mechanisms underlying species-specific methylation differences; (3) identifying cis-acting genetic variants and trans-acting factors underlying these differences; (4) identifying lineage-specific natural selection on methylation; and (5) linking this selection to gene expression and candidate traits. Addressing these gaps would enhance our understanding of how genetic variants impact methylation and in turn shape adaptive traits.

Human-chimpanzee tetraploid hybrid induced pluripotent stem cells (iPSCs), generated through in vitro cell fusion, provide a controlled experimental system for quantifying cis– and trans-regulatory contributions to methylation divergence by hosting both species’ alleles in an identical trans-environment. By measuring ASM and ASE across multiple cell types, this system generates genome-wide maps of cell type-specific regulatory differences that may contribute to evolutionary adaptation. In this study, we generated bisulfite sequencing data from human, chimpanzee, and hybrid iPSCs, cranial neural crest cells, dopaminergic neurons, skeletal myocytes, and hepatocytes, as well as from primary human and chimpanzee dental pulp stem cells. We combined these data with previously published26 and newly generated RNA-seq data from all six cell types.

Using these data, we investigated the causes of methylation divergence and quantified contributions of cis– and trans-acting mechanisms that establish interspecies differences in methylation patterns. We developed a hypothesis testing framework tailored to count-based proportional methylation data and extended a changepoint detection algorithm to identify differentially methylated regions (DMRs) under consistent cis– or trans-regulation. We show that cis-acting mechanisms predominantly drive genome-wide methylation divergence, with cis-regulated DMRs likely affected by nearby species-specific single-nucleotide variants. In contrast, trans-regulated DMRs were enriched for transcription factor binding motifs, including pioneer factors and chromatin remodelers that may mediate trans-acting divergence. Concordance between ASM and ASE across cell types revealed species-specific and context-dependent regulatory relationships linking methylation to gene expression. Finally, coordinated methylation-expression divergence in gene sets related to speech, cognition, and viral infection susceptibility connected lineage-specific selection to candidate phenotypes. In sum, our work provides a framework for understanding how divergent cis– and trans-acting factors converge on methylation patterns to produce species-specific traits.

Results

Measurement of DNA methylation across diverse cell types

To establish comprehensive methylation landscapes across evolutionarily relevant cell types, we differentiated iPSCs from two humans, two chimpanzees, and two fused tetraploid hybrids into four cell types (see Methods)24,26. We included an additional cell type, dental pulp stem cells, which are primary cells isolated from two human and two chimpanzee donors24. We then generated whole-genome bisulfite sequencing (WGBS) data from human and chimpanzee parental and hybrid cranial neural crest cells (CNCC) and dopaminergic neurons, as well as parental dental pulp stem cells (DPSC), complemented by reduced representation bisulfite sequencing (RRBS) data from induced pluripotent stem cells (iPSCs), skeletal myocytes (SKM) and hepatocytes (HEP) (Figure 1A). For cross-cell-type analyses combining WGBS and RRBS data, we restricted to CpG sites covered in all samples after phasing, totaling 561,762 shared CpG sites. For hybrid samples, reads were assigned to the human or chimpanzee allele using species-specific SNV positions; on average, 51% ± 0.4% of all aligned reads were allele-resolved across hybrid WGBS samples (DA and CNCC), and 10.6% ± 0.2% of aligned reads were allele-resolved across hybrid RRBS samples (IPSC, SKM and HEP). The lower rate in RRBS reflects its short read lengths and its enrichment for CpG-dense promoter regions, which are evolutionarily conserved and therefore depleted of informative human–chimpanzee SNVs (see “RNA-sequencing and allele-specific expression analysis” in Methods section). Principal component analysis and hierarchical clustering of methylation patterns showed that samples clustered primarily by cell type, then by species, and finally by parental (human and chimpanzee) versus hybrid system (Figure 1B, 1C and Figure 1—figure supplement 1). CNCC samples displayed a unique clustering pattern where parental vs. hybrid samples clustered separately, though examination of the dendrogram indicated this was due to a single human parental sample being a slight outlier (Figure 1B). The transcriptional distinctiveness of DPSCs relative to iPSC-derived cell types likely reflects their status as primary cells, which retain the epigenetic and transcriptional signatures of their tissue of origin and have not undergone reprogramming (Figure 1C). Known cell type-specific gene expression markers validated the correct identity of each differentiated cell type (Figure 1D and Figure 1—figure supplement 2)35–49. Overall, these results are consistent with our expectation that cell type is the predominant determinant of DNA methylation patterns, while species explains a much smaller proportion7. In other words, a human hepatocyte is much more similar to a chimpanzee hepatocyte than it is to a human dopaminergic neuron.

DNA methylation profiles across diverse human, chimpanzee, and hybrid cell types.

(A) Four cell types were differentiated from human, chimpanzee, and hybrid induced pluripotent stem cells. The cell types span diverse body systems including skeletal myocytes for the musculoskeletal system, dopaminergic neurons for the central nervous system, cranial neural crest cells for craniofacial development and hepatocyte progenitors for the liver. We collected DNA methylation and RNA-seq data from these cell types from humans, chimpanzees and their hybrids. Dental pulp stem cells were also collected from human and chimpanzee adult tissues. (B) Clustering analysis and (C) PCA between samples for all cell types (left) and for all iPSC-derived cell types (i.e. excluding DPSC, right). Parent/hybrid stratifications are shown separately in Figure 1—figure supplement 1. Each replicate represents an independent biological differentiation from the indicated iPSC line. (D) Cell type specific markers of gene expression. Figure 1A was created using BioRender.

Contribution of cis and trans regulation to DNA methylation divergence between human and chimpanzee

To understand the molecular mechanisms underlying species-specific methylation patterns, we first established a framework for distinguishing between cis– and trans-regulatory control of DNA methylation. By comparing methylation levels between human and chimpanzee alleles within the same cellular environment (hybrid cells), we can isolate cis effects from trans effects and determine which mechanism predominates in establishing species-specific methylation patterns (Figure 2A).

Contribution of cis and trans regulation to DNA methylation divergence between human and chimpanzee.

(A) Interspecies differences in methylation levels per CpG site is compared between parental and hybrid systems and assigned a regulation category based on hypothesis testing. (B) Parental and hybrid methylation ratios in cranial neural crest cell promoters and the respective regulation category at FDR < 0.05 (color key in panel C). Parental and hybrid methylation differences are plotted as the per-sample average difference of fractional methylation. (C) Quantification of relative contribution of non-conserved regulation groups to whole-genome CpG sites across cell types at FDR < 0.05. (D) Heterogeneity of the distribution of CpG sites of all regulation groups. (E) Quantification of CpG regulatory clusters by overlap with genomic features of transcripts genome-wide as well as cis-regulatory elements including promoters, enhancers and CTCF-bound cis-regulatory regions (CTCF). (F) Quantification of relative contribution of non-conserved regulation groups to regulatory clusters across cell types (color key in panel C). Figure 2A was created using BioRender.

To quantify the relative contributions of cis-versus trans-acting factors in establishing human-chimpanzee methylation differences, we applied a beta-binomial model framework adapted from Hallgrímsdóttir et al. (2024)50 to classify differentially methylated regions based on allele-specific patterns in hybrid versus parental samples. To determine whether a CpG site reflects a difference in methylation due to cis or trans regulation, two hypothesis tests are used: one to evaluate the null hypothesis that the difference in methylation is due solely to trans, and the other solely to cis (see Methods). Each site is assigned to one of five mutually exclusive regulatory categories. Cis-only sites show allelic differences in hybrids that are also preserved in the parental comparison. Trans-only sites show allelic differences in parentals that are absent in hybrids. Cis + trans (additive) sites show allelic differences in both systems, with the same species methylated higher in both, but with the magnitude of interspecies difference reduced in hybrids, indicating a partial trans contribution acting in the same direction as cis. Cis × trans sites show allelic differences in both systems but with opposite directionality (one species higher in parentals, the other higher in hybrids). Conserved sites show no significant allelic differences in either system (Figure 2A and 2B). We investigated genome-wide methylation divergence by fitting the model to methylation levels at each CpG site (totaling 49,049,052 CpG sites across all cell types; see Supplemental File 1 for detailed statistics). In all cell types examined, most CpG sites (83–93%) had conserved patterns of methylation. Among divergent sites, nominally-classified cis-regulation was over twice as common as trans-regulation (Figure 2—figure supplement 1, Supplemental File 1). At FDR < 0.05, cis-regulation accounted for a mean of 91% of divergent sites and trans-regulation a mean of 4.5% across cell types, and cis remained the dominant category at every FDR cutoff tested (FDR < 0.25–0.01; Figure 2C, Figure 2—figure supplement 1–2, Supplemental Files 1–2). The genome-wide predominance of cis-regulation was further supported at the regional level (Figure 2B and Figure 2—figure supplement 1 and 3). This genome-wide predominance of cis-regulation indicates that local sequence differences between human and chimpanzee genomes are the primary drivers of methylation divergence, consistent with previous hypotheses of gene regulation that cis-regulatory changes may be less disruptive than broad trans-acting modifications1–4,6.

The regional distribution of sites from different regulation groups reveals a consistent pattern across cell types: conserved CpG sites are most likely to be near other conserved sites, while divergent CpG sites are likely to be near conserved sites as well as sites within the same regulation group (Figure 2D). This suggests that CpG sites within the same regulation group tend to form coherent clusters in the genome, and divergent CpG clusters typically have conserved sites scattered within. When conducting regional analyses, methylation data is usually pooled from multiple sites within a pre-defined genomic region6–10. However, this regional heterogeneity can affect classification as well as estimates of how much each regulatory mechanism contributes to divergence between species (Figure 2—figure supplement 2-3, Supplemental File 2, see Methods). Therefore, pooling data across regions should be approached carefully, as subsets of CpG sites within the same regulatory element can be controlled by distinct mechanisms. These findings underscore the importance of studying methylation divergence at high resolution.

To identify differentially methylated regions (DMRs) exhibiting consistent inter-species methylation directionality and minimal classification heterogeneity of regulation groups, we implemented a changepoint detection method with a tolerance threshold for regulatory homogeneity (see Methods). This method identifies genomic loci where co-localized CpG sites demonstrate uniform regulatory behavior punctuated by a pre-defined fraction of heterogeneous sites: non-conserved classifications showing consistent directional methylation bias, and conserved sites maintaining equal methylation levels between species. This method prioritizes DMRs that are controlled by the same mechanism and display coordinated methylation patterns (Supplemental File 3), thereby enriching for loci with interpretable regulatory mechanisms. Notably, DMRs identified through this method are enriched in 5’ UTRs (just downstream of each transcription start site [TSS]) and CTCF-bound regions, supporting their functional relevance in gene regulation (Figure 2E). In total, we identified 9159 cis-DMRs and 2046 trans-DMRs across all cell types (see Methods and Supplemental File 3). The quantification of cis– vs. trans-DMRs yielded the same conclusion that cis-regulation outweighs trans– regulation in DNA methylation divergence (Figure 2F).

The causes of cis– and trans-acting divergence in DNA methylation

To identify the molecular mechanisms driving methylation divergence, we analyzed transcription factors (TFs) associated with trans-DMRs and sequence variants associated with cis-DMRs.

TFs represent a major class of trans-acting factors capable of establishing and maintaining methylation patterns across the genome, often by recruiting methyltransferase or demethylase enzymes. Interspecies variation in TF expression levels, binding activity, or target site accessibility can propagate to downstream methylation differences at their binding sites. To identify TFs potentially driving species-specific methylation divergence through such trans-acting mechanisms, we performed TF binding site motif enrichment analysis on trans-DMRs (Figure 3A, Supplemental File 4 and 5)51. This analysis identified several candidate factors, including a pioneer factor FOXA2 (Benjamini-Hochberg corrected FDR = 3.7 × 10⁻⁵ in DA Hu<Ch trans-DMRs) and a chromatin modifier FOXM1 (FDR = 0.002, 4.8 × 10⁻⁶, and 0.021 in CNCC, DA, and HEP Hu<Ch trans-DMRs, respectively), both known to actively reshape DNA methylation landscapes52–58.

If species-specific levels of a TF are responsible for the enrichment of its predicted binding sites in trans-DMRs, then this enrichment may be strongest in cell types with greater species-specific differential expression of the TF. FOXM1 and FOXA2 each showed a pattern of this kind, with higher relative expression in one species accompanied by lower methylation at their predicted binding sites in that same species and cell type. FOXM1 was over 4-fold more highly expressed in human HEPs than in chimpanzee, and its predicted binding sites were highly enriched specifically for trans-DMRs with lower methylation in human HEPs; the three other cell types, in which human and chimpanzee expression of FOXM1 is more similar, showed weaker enrichment and little species-specific asymmetry in trans-DMR methylation directionality (Figure 3B). We observed a similar pattern for FOXA2 (Figure 3—figure supplement 1). We note that with only four cell types available per comparison, we are not adequately powered to test for a statistically significant relationship between differential TF expression and motif enrichment. We therefore present FOXM1 and FOXA2 as exploratory, hypothesis-generating examples rather than as evidence of a genome-wide pattern.

Cis– and trans-factors influencing species-specific methylation patterns.

(A) HOMER motif enrichment of trans-DMRs in regions Hu>Ch and Ch>Hu methylated in humans (left panel), and differential expression of corresponding transcription factors in parents (right panel). (B) Fold-enrichments against genomic background of FOXM1, one of the top motifs enriched in Hu>Ch and Ch>Hu trans-DMRs, plotted against Hu and Ch relative gene expression levels in parents. (C) Three classes of SNVs—CpG gains in humans, CpG losses in humans, or no CpG change—and their effects on methylation of conserved neighboring CpG sites within ±25, 50, and 100 bp. The differential methylation patterns around CpG SNVs versus non-CpG-SNVs are assessed using two-proportion test (***: p < 0.001, **: p < 0.01, *: p < 0.05) (D) Distances of each cis-DMR vs trans-DMR to the nearest CpG SNV site (Mann-Whitney U test; ***: p < 0.001).

Other TFs showed different patterns of association between species-specific expression and trans-DMR methylation. For example, FOXP1 and OTX2 were most differentially expressed in HEPs, and their predicted binding sites were also most enriched in trans-DMRs in this same cell type (Figure 3—figure supplement 1). However, unlike FOXM1 and FOXA2, binding sites for FOXP1 and OTX2 showed similar patterns of enrichment across cell types for both Hu>Ch and Ch>Hu methylated trans-DMRs. This lack of directionality may suggest that these transcription factors are not restricted to either demethylation or methylation, perhaps having a more indirect or context-specific role in regulating the epigenetic landscape59–61, though further work is required to test this hypothesis. Altogether, our trans- DMR analysis nominated many TFs that may contribute to methylation divergence, including several with trans-DMR binding site motif enrichment associated with the TF’s differential expression that we highlight for future experimental testing.

To identify the molecular mechanisms driving cis-regulatory methylation differences, we then examined whether local sequence variants could explain cis-acting methylation divergence. CpG dinucleotides are the primary targets of DNA methylation in mammalian genomes, and their presence, density, and spacing are known to influence local methylation patterns62–64. Therefore, single nucleotide variants (SNVs) that either create or disrupt CpG dinucleotides represent a plausible mechanism by which sequence divergence could drive species-specific methylation differences. To investigate this hypothesis, we systematically analyzed SNVs and their associated methylation differences in neighboring conserved CpGs within windows of 25, 50, and 100 base pairs upstream and downstream of each SNV. Our analysis revealed that SNVs affecting CpG dinucleotides—but not other SNVs—are associated with altered methylation patterns of nearby CpG sites up to 50 base pairs from the variant site (Figure 3C). Notably, the effect diminished with increasing distance from the SNV, with the strongest methylation changes observed within 25 base pairs and progressively weaker effects at 50 base pairs. Beyond 50 base pairs, the magnitude of influence of CpG SNVs on neighboring methylation patterns became negligible, suggesting a limited spatial range for these local sequence effects. This distance-dependent relationship indicates that sequence variants affecting CpG density and spacing might influence the local recruitment or activity of DNA methyltransferases or demethylases. We note that CpG SNVs represent just one of many possible mechanisms for cis-methylation divergence; other classes of variants, such as those disrupting transcription factor binding sites, may also contribute and warrant future investigation.

This model predicts that while cis-DMRs should often be close to CpG SNVs, trans- DMRs—which arise from regulatory mechanisms independent of nearby sequence variation— should not be. To test this, we compared the genomic distance between each DMR and its nearest CpG SNV. Consistent with our model, we found consistently shorter distances for cis- DMRs than trans-DMRs across all cell types examined: Median distances to the nearest CpG SNV were 24 ± 6 bp for trans-DMRs vs. 11 ± 2 bp for cis-DMRs (Figure 3D; Mann-Whitney p < 0.001 for all comparisons)65. This difference suggests that cis-DMRs tend to occur near CpG sequence variants and are likely directly influenced by local changes in CpG content, while trans-DMRs show no preferential association with CpG variants, as expected given that trans-DMRs are by definition not driven by local sequence variation.

To further test the effects of CpG SNVs on local methylation, we assessed whether the same SNVs show consistent methylation effects across multiple cell types. CpG SNVs associated with a cis-DMR in one cell type were significantly more likely to be associated with a cis-DMR in other cell types than expected by chance, consistent with a shared mechanistic basis for cis-methylation divergence across cell types (Figure 3—figure supplement 2). This genomic organization is consistent with trans-DMRs being regulated by diffusible factors such as transcription factors rather than by local sequence variants, supporting a model in which cis– and trans-acting mechanisms operate through fundamentally different molecular pathways to generate species-specific methylation patterns.

Cell type– and species-specific gene expression as a consequence of methylation divergence

Given the well-established repressive effect of promoter methylation on gene expression, we assessed whether methylation differences contribute to expression divergence by analyzing genome-wide correlations between allele-specific promoter methylation and expression levels across different genomic contexts (Figure 4A)66–68. We first examined whether cis-acting variants drive promoter methylation consistently across cell types or in a context-dependent manner. We found that while some promoters are allele-specifically methylated across all cell types, many more showed cell type-specific cis-methylation patterns, suggesting that promoter methylation regulation is often context-dependent (Figure 4B and 4C). However, while DMRs consistently detected across cell types provide strong evidence of shared regulation, DMRs detected in only one cell type may reflect either genuine cell type-specificity or insufficient power to detect them elsewhere. Our estimate of cell type-specific regulation therefore represents an upper bound on the degree of cell type-specific methylation regulation.

Human-chimpanzee allele-specific methylation can be associated with allele-specific expression.

(A) Illustration of the transcriptional consequences of DNA methylation explored in this study. DNA methylation, as a stable cis-acting regulatory mark, often represses transcription. In a hybrid, allele-specific methylation can lead to allele-specific gene expression, although both methylation and its effects on transcription can be context-dependent. (B) We identified many promoters with allele-specific methylation in each cell type, most of which showed allele-specific methylation in only one cell type. (C) Quantification of promoters with pure-cis regulation of methylation across cell types. (D, E, F) Examples of gene expression-methylation correspondence. Methylation tracks show fractional methylation (mC/coverage × 100%) of individual CpGs. Expression values are in TMM (Trimmed Mean of M-values)-normalized CPM (counts per million). (D) CTSF as an example of species-specific, cell type-agnostic DMR, (E) LGALS8 as an example of cell type-specific allele-specific methylation, and (F) HOXA9 as an example of cell type-specific but species-agnostic methylation patterns. (G) Genome-wide profile of expression-methylation correlations (binned by distance of methylated region to the TSS) for genes with both promoter methylation and expression under cis-regulation (dark red and dark blue) or trans-regulation (light brown and light gray). Detailed statistics are provided in Supplemental File 6. Figure 4A was created using BioRender.

Several specific examples illustrate different patterns of methylation-expression coordination (Supplemental Files 5 and 6; see Methods). The promoter of CTSF is a cell-type agnostic cis-DMR with consistent Hu>Ch methylation across cell types corresponding to coordinated Ch>Hu expression (Figure 4D and Figure 4—figure supplement 1). In contrast, a putative enhancer in LGALS8 represents a case where methylation divergence is confined to specific cellular contexts. The candidate enhancer overlaps with a cis-DMR with Hu>Ch methylation and LGALS8 exhibits allele-specific Ch>Hu expression exclusively in CNCCs, while having no divergence in methylation or expression in other cell types (Figure 4E and Figure 4—figure supplement 2). A third type of pattern is exemplified by the HOXA9 promoter, with cell type-specific hypomethylation corresponding to cell type-specific expression only in SKM (Figure 4F). These examples illustrate how different regulatory mechanisms can produce distinct patterns of methylation-expression coordination and how epigenetic changes may contribute to both species-specific and cell type-specific regulatory evolution.

Previous studies have reported weak but significant correlations between DNA methylation and gene expression divergence in primates7,12,14,16,69. Consistent with these findings, our genome-wide estimates of human–chimpanzee transcriptome divergence produced similar weak but significant correlations with methylation divergence (Figure 4—figure supplement 3 and Supplemental File 6). Our comparison of CpG island (CGI) vs. non-CGI promoters also yielded trends consistent with past studies showing that CGI promoter methylation correlates more strongly with gene expression repression than does non-CGI promoter methylation70,71 (Figure 4—figure supplement 4).

However, such genome-wide correlations may mask stronger relationships that occur within specific regulatory contexts where methylation and expression are more directly coupled. We hypothesized that our hybrid system could help identify genes exhibiting stronger methylation–expression coupling by stratifying them based on regulatory mechanisms. We reasoned that genes whose methylation and expression are both influenced by cis-acting variants— potentially driven by the same underlying genetic differences— should display tighter coupling than genes regulated through independent mechanisms.

To test this, we evaluated the extent to which methylation and expression show a negative relationship (i.e., higher methylation associated with lower expression) consistent with a repressive effect of methylation. In both CNCC and DA hybrid cells, we found that genes with cis-regulated expression were more likely to exhibit repressive patterns than those with trans-regulated expression (Figure 4—figure supplement 5). This is consistent with DNA methylation acting as an upstream cis-regulatory mechanism: genes with stronger cis-regulation are expected to show tighter methylation-expression coupling due to the greater contribution of local regulatory factors, including methylation itself. Further restricting the analysis to genes with concordant regulatory mechanisms revealed that genes with both methylation and expression being cis-regulated (“cis–cis” genes; Supplemental File 8) exhibited significantly stronger negative methylation–expression correlations than “trans–trans” genes (Figure 4G, see Supplemental File 6 for detailed statistics). This supports the hypothesis that methylation– expression coupling is strongest when both are driven by local genetic variation, perhaps through shared causal variants. Different shapes of the methylation–expression correlation profiles around the TSS in each cell type may reflect differences in the chromatin landscape and transcription factor occupancy that modulate the functional impact of promoter methylation in each cell type context; a deeper mechanistic explanation is left for future work.

The sign test identifies lineage-specific selection on DNA methylation between human and chimpanzee

To identify pathways that have evolved under lineage-specific selection, we used a two-step binomial sign test approach that tests for non-random directional bias in methylation-expression patterns in hybrids1,23,24,26,72–76. The use of hybrid cell lines is critical for this analysis as it ensures that different promoters represent independent observations, controlling for trans-acting factors and environmental variation that could otherwise confound the detection of cis-regulatory effects. This test operates on the logic that neutral evolution should act like flipping a fair coin: if human-biased and chimpanzee-biased changes occur at equal background frequencies, then methylation and expression changes of genes within a pathway would accumulate randomly, showing no preference for human-biased versus chimpanzee-biased directionality. In contrast, lineage-specific selection would bias the coin flip, potentially causing functionally related gene sets to consistently favor one direction over the other rather than splitting evenly by chance.

To maximize our ability to find phenotypically impactful divergence, we adapted the standard sign test to account for directionality of both methylation and gene expression. In the first step, we applied a standard binomial sign test to all 6,283 promoters showing ASM, counting the number of genes showing human-biased versus chimpanzee-biased methylation changes within each gene set. The background probability for this test was defined as the overall fraction of promoter ASM in each cell type that was human-biased. This initial analysis identified gene sets showing significant enrichment for directional methylation changes, establishing candidates for lineage-specific regulatory divergence. In the second step, we restricted our attention to genes showing a negative association between ASM and ASE. This subset represents cases where methylation changes are most likely to have a simple and direct repressive effect on transcription, filtering out more complex regulatory scenarios where methylation and expression changes may be uncoupled or subject to competing regulatory influences.

Gene sets showing significant deviation from neutral expectation in the ASM/ASE concordance test (nominal binomial p-value < 0.05 and FDR < 0.25 after multiple test correction; see Methods), with supporting evidence of directional bias in the methylation-only sign test (nominal binomial p-value < 0.05), are interpreted as likely targets of lineage-specific selection on promoter methylation that could directly impact gene expression (Figure 5, Supplemental File 7). However, as indicated by the FDR, we do expect some fraction of these candidate gene sets to be false positives. Compared to a standard sign test that focuses solely on directionally biased gene expression72, this more stringent two-step filtering approach may identify higher-confidence candidates where regulatory divergence is reflected at both the methylation and expression levels, establishing a potential mechanistic connection to phenotype.

Gene sets with evidence of lineage-specific selection on methylation and gene expression.

The length of the bars indicates the number of genes in each gene set with allele-specific expression or methylation in each direction (human or chimpanzee-biased). The bars in darker colors represent the number of genes where higher methylation is associated with repressed gene expression, whereas lighter colors represent genes where higher methylation is associated with increased gene expression.

Among gene sets from Kyoto Encyclopedia of Genes and Genomes (KEGG) and Human Phenotype Ontology (HPO)77,78, we identified several with clear relevance to human-specific traits. In iPSC hybrids, multiple gene sets including “poor speech,” “highly arched eyebrow,” and “growth delay” all showed Hu>Ch methylation and Ch>Hu expression (Figure 5, Figure 6A and Figure 6B). In the SKM hybrids, genes related to “inability to walk” showed significant Hu>Ch methylation and Hu<Ch expression (Figure 5). In DA hybrids, genes involved in “intellectual disability, moderate” show significant Hu <CH methylation and Hu> Ch expression (Figure 5 and Figure 6C). In HEP hybrids, we identified the “Hepatitis C” gene set with significant Hu<Ch methylation and Hu>Ch expression (Figure 5 and Figure 6D). In primary DPSC samples, we identified significant bias toward Hu<CH methylation and Hu>Ch expression in genes related to tooth eruption timing, consistent with known differences in dental development between humans and chimpanzees, where human permanent molars emerge approximately 6 years apart compared to 3-4 years in chimpanzees (Figure 5—figure supplement 1)79–81. However, the DPSC sign test results should not be interpreted as evidence of lineage-specific selection due to the use of parental rather than hybrid cell lines, which may introduce dependencies among observations due to trans-acting effects, potentially inflating statistical significance. For all other cell types (iPSC, DA, SKM, HEP, CNCC), the sign test results are based on hybrid lines and can be interpreted as evidence of lineage-specific selection, as the hybrid system controls for trans-acting confounders.

Directional bias in allele-specific methylation and gene expression identifies candidate genes for human-specific phenotypes.

Allele-specific expression (ASE) is plotted against allele-specific methylation (ASM). Orange dots represent genes consistent with repressive methylation (e.g. Hu>Ch ASE and Hu<Ch ASM) and purple dots represent genes consistent with activating methylation (e.g. Hu<Ch ASE and Hu>Ch ASM). Bar plots represent counts of genes with repressive methylation that exhibit expression changes in directions either consistent with (light green) or opposite to (light pink) what would be expected based on phenotypic differences between humans and chimpanzees. A) The “highly arched eyebrow” gene set in iPSC hybrids shows consistent Hu<CH ASE and Hu>Ch ASM for genes with repressive methylation. B) The “growth delay” gene set in iPSC hybrids shows significant bias towards Hu<CH ASE and Hu>Ch ASM for genes with repressive methylation. C) The “intellectual disability, moderate” gene set in DA shows consistent Hu>Ch ASE and Hu<Ch ASM for genes with repressive methylation, with genes including GRIK2, TUBB3, EGF and CC2D2A being among the genes with highest magnitude ASE and ASM. D) The “Hepatitis C” gene set in HEP shows consistent Hu<CH ASE and Hu>Ch ASM for genes with repressive methylation. E) The “poor speech” gene set in iPSC hybrids represents a potentially compensatory pathway where most genes with repressive methylation show Hu<CH ASE and Hu>Ch ASM, whereas the genes with highest magnitude of ASE and ASM (TUBB3 and GRIK2) show Hu>Ch ASE and Hu<Ch ASM.

We then analyzed individual genes within each candidate gene set to assess whether expression bias in associated phenotypes aligns with known human-chimpanzee trait differences, focusing on genes showing repressive methylation. Multiple gene sets demonstrated directional bias in gene expression consistent with the direction of phenotypic divergence found between humans and chimpanzees. For instance, in the “highly arched eyebrow” gene set—where most genes show human down-regulation in iPSC hybrids— COLEC11 (which has 8-fold lower expression from the human allele) stands out as an outlier in magnitude. COLEC11 acts as a guidance cue in directing neural crest cell migration during early development, and in humans its loss-of-function (LOF) causes craniofacial dysmorphism including highly arched eyebrows and underdeveloped brow ridge, both of which are traits unique to humans (Figure 5 and Figure 6A)82. In the “growth delay” pathway, most genes with repressive methylation (whose LOF in humans and other species leads to delayed growth) show lower expression from their human alleles, consistent with humans’ substantially delayed growth pattern compared to chimpanzees (Figure 6B)83,84. Many of the genes in DA whose LOF causes “intellectual disability, moderate” may have contributed to the evolution of human cognitive abilities. For example, the high-magnitude genes with Hu>Ch ASE and Hu<Ch ASM such as GRIK2, TUBB3, EGF and CC2D2A are essential for excitatory synaptic plasticity and long-range axon wiring (Figure 6C)85–90. In addition, most genes whose inhibition or LOF increases susceptibility or severity of Hepatitis C (e.g. MX1 and OAS3) show lower expression from human alleles in HEP, consistent with human’s heightened vulnerability to Hepatitis C viral infections (Figure 6D)91–93.

Interestingly, we also discovered gene sets where the direction of most gene expression differences predicted the opposite of a known human/chimpanzee phenotypic difference. For example, in the “poor speech” gene set, where loss-of-function is associated with impaired speech in humans, we counterintuitively observed that most genes were down-regulated in humans compared to chimpanzees. However, the two genes with the largest fold-changes, TUBB3 and GRIK2, showed higher expression from the human allele (Figure 6E). Notably, these two genes are also among the large-magnitude genes in the predominantly human up-regulated “intellectual disability, moderate” pathway, which does align with species-specific phenotype (Figure 6C). This suggests the speculative possibility of compensatory evolution in which large, pleiotropic effects of a few genes are counterbalanced by smaller changes in many genes (Figure 6C and Figure 6E; also see Discussion). Our previous study found that genes affecting voice box show particularly extensive down-regulation in humans, consistent with our observation in the “poor speech” pathway24.

Collectively, these results provide the first evidence for lineage-specific selection on DNA methylation. Our use of hybrid cell lines was critical to this discovery, as it allowed us to isolate cis-acting regulatory changes from confounding trans-acting and environmental effects. Isolating cis-acting changes is critical because the sign test requires that each promoter represents an independent observation; in hybrid cells, each allele’s methylation is determined by its own cis-regulatory sequence, making observations across promoters genuinely independent and thus satisfying the assumptions of the binomial test. This approach reveals that polygenic selection has acted on entire sets of functionally related genes whose expression is modulated through DNA methylation. Importantly, by connecting lineage-specific methylation divergence to expression changes within functionally coherent pathways, we can generate plausible hypotheses linking regulatory evolution to human-specific phenotypes. These include craniofacial morphology, dental development, viral infection susceptibility, and neurodevelopmental characteristics underlying speech and cognition (Figures 5 and 6, Figure 5—figure supplement 1).

Discussion

Our investigation using human-chimpanzee hybrid tetraploid cells revealed that cis-regulatory mechanisms are the primary drivers of genome-wide methylation divergence between humans and chimpanzees across multiple cell types, including dopaminergic neurons, skeletal muscle, iPSCs, hepatocytes, and cranial neural crest cells. This finding aligns with previous studies highlighting the predominant role of cis-regulatory elements in evolutionary adaptations, as they enable precise, context-specific gene regulation while minimizing the potentially deleterious pleiotropic effects of trans-acting factors6. We demonstrated that most differentially methylated sites arise through cis-regulatory mechanisms, with trans-regulation and cis-trans interactions playing secondary roles. We identified clusters of CpG sites with consistent regulatory mechanisms, forming coherent differentially methylated regions (DMRs). These DMRs predominantly overlap with functionally relevant genomic features including 5’ UTRs and upstream promoter regions, underscoring their regulatory significance.

Our mechanistic analysis revealed distinct pathways through which cis– and trans-regulatory factors establish species-specific methylation patterns. For cis-regulation, we identified CpG-altering sequence variants as a direct mechanism linking genomic divergence to epigenetic differences. These variants influence methylation patterns of neighboring CpG sites with effects diminishing with genomic distance. Importantly, cis-DMR clusters are significantly closer to CpG SNVs than trans-DMRs across all cell types, providing evidence that local sequence variation directly shapes cis-regulatory methylation patterns, perhaps through altered CpG density affecting recruitment of DNA methyltransferases or demethylases.

Our analysis of allele-specifically methylated promoters revealed that the majority showed cell type-specific patterns. Our examples illustrate this diversity: the CTSF promoter exhibits consistent cross-cell-type cis-methylation divergence (Figure 4—figure supplement 1) with corresponding expression changes; an enhancer near LGALS8 shows cell type-specific allele-specific methylation and overlaps with a homogeneous cis-DMR exclusively in cranial neural crest cells; and the HOXA9 promoter displays cell type-specific methylation patterns with corresponding context-dependent expression effects. These findings highlight how the broader regulatory context of each cell type impacts the effects of cis-regulatory variants.

Our genome-wide correlation analyses confirmed the well-known pattern that while promoter methylation generally shows statistically significant inverse relationships with gene expression, these correlations are weak when examined across all genes. However, by leveraging our hybrid cell system to stratify genes based on regulatory mechanism, we uncovered significantly stronger negative correlations between methylation and expression when both are cis-regulated compared to genome-wide estimates or trans-regulated genes. This finding suggests that the repressive effects of species-specific methylation are strongest when both layers of regulation are driven in cis, perhaps because cis-regulated methylation changes tend to have larger effect sizes than trans-regulated changes.

Our two-step binomial sign test approach in hybrids identified multiple gene sets showing coordinated directional shifts in methylation and expression that significantly exceed neutral expectations. Many of the pathways we identified are related to known human-chimpanzee phenotypic differences: intellectual disability in hybrid DAs, growth delay and highly arched eyebrow in hybrid iPSCs, hepatitis C in hybrid HEPs, and tooth eruption timing in parental DPSCs (Figure 5, 6, Figure 5—figure supplement 1). Specifically, the evidence rejecting a neutral null model of cis-regulatory evolution consists of statistically significant directional bias in both promoter methylation and gene expression within functionally coherent gene sets, assessed using a two-step binomial sign test with permutation-based FDR correction. These findings represent the first evidence for coordinated directional divergence in DNA methylation and identify candidate genes under lineage-specific selection. By restricting analysis to genes showing negative methylation-expression associations—where methylation most likely exerts direct transcriptional repression—we identified high-confidence candidates where epigenetic divergence has likely functional consequences. Our results suggest that cis-regulatory changes in DNA methylation represent an important substrate for adaptive evolution, complementing protein-coding and other cis-regulatory changes in generating human-specific traits.

We developed several methodological innovations tailored to the unique challenges of comparative epigenomics. First, our beta-binomial framework accounts for the count-based nature of bisulfite sequencing data and explicitly models allele-specific patterns in hybrid versus parental contexts. Second, our changepoint detection algorithm identifies coherent DMRs with minimal regulatory heterogeneity, prioritizing functionally organized regulatory units over isolated CpG sites. Third, the two-step sign test we introduce provides a method to identify lineage-specific selection that links methylation and corresponding changes in gene expression. These methods can be readily applied to other taxa where hybrid organisms can be generated.

However, our study also has important limitations. First, while hybrid cells provide powerful controls for trans-acting variation, in vitro differentiation protocols do not fully recapitulate the complexity and temporal dynamics of in vivo development. Therefore, cell lines are limited in their ability to capture species-specific methylation or expression profiles that are specific to a certain developmental stage or in vivo context. Second, pathway analysis of DPSCs used primary human and chimpanzee samples rather than hybrids, with trans-acting factors potentially inflating statistical significance in the sign test; these results should therefore be interpreted not as evidence for positive selection, but rather as an interesting pattern that could have important connections to phenotypic differences between humans and chimps. Third, while we identified associations between transcription factor binding motifs and trans-DMRs, definitive demonstration of causality requires targeted experimental perturbations such as base editing (for CpG-disrupting SNVs) or TF knockdown/overexpression experiments (for trans-acting TFs). Fourth, our study examined a limited number of cell types, and methylation divergence in most tissues remains unexplored. Fifth, our study lacks an outgroup species; therefore, we cannot definitively attribute observed differences to the human lineage specifically, and our use of the term “human-specific” refers to differences between humans and chimpanzees rather than changes unique to the human lineage. Sixth, while we focused on CpG SNVs as a mechanism for cis-methylation divergence, other classes of variants—such as transcription factor binding site mutations—could also influence local methylation patterns and represent an important avenue for future investigation.

Comparative gene expression analyses between humans and closely related species offer a powerful way to explore the molecular basis of species-specific traits and disease susceptibilities. However, it remains challenging to infer phenotypic differences through the lens of human clinical ontologies. To start with, HPO terms are largely derived from pathological human data and predominantly capture phenotypic consequences of loss-of-function mutations in a clinical context. When comparing expression levels between species, one might assume that lower expression corresponds to a loss-of-function-like phenotype. However, the relationship between expression dosage and function is typically nonlinear94, and complete gene knockouts in a model organism or loss of function mutations in patients are not equivalent to moderate downregulation. This fact may help explain cases where our sign test results suggest the opposite directionality as compared to human/chimpanzee phenotypic differences (Figure 6E). Alternatively, humans and chimpanzees may achieve a similar phenotype through distinct configurations of regulatory networks. For example, an adaptive down-regulation of a particular gene may be compensated for by up-regulation of other genes in the same pathway, maintaining overall pathway function. In other words, compensatory changes may complicate the relationship between gene expression and phenotype, as may be the case for the “poor speech” gene set (Figure 6E)23.

Several important directions emerge from this work. First, extending this framework to additional cell types and developmental time points would reveal whether the predominance of cis-regulation we observe is universal or varies across developmental contexts. Longitudinal sampling during differentiation could identify critical windows when methylation patterns are established and reveal whether human-chimpanzee differences emerge gradually or through discrete developmental transitions. Second, integrating chromatin accessibility (ATAC-seq) and histone modification (ChIP-seq) data would provide a more complete picture of the regulatory landscape, clarifying how DNA methylation interacts with other chromatin features to establish gene expression patterns. Third, direct examination of transcription factor binding using techniques like CUT&RUN in both species could validate the predicted relationships between transcription factor occupancy and methylation patterns at trans-DMRs. Finally, extending these analyses to include other great apes as an outgroup would provide phylogenetic resolution to distinguish changes that occurred in the human vs. the chimpanzee lineage.

Collectively, our findings provide a framework for understanding how DNA methylation divergence contributes to human-specific traits. We demonstrate that while both cis– and trans-regulatory mechanisms shape interspecies methylation differences, cis-acting factors predominate, thus directly linking genomic sequence variation and epigenetic divergence. The coordinated nature of many independent methylation-expression changes across phenotypically relevant pathways is consistent with epigenetic modifications being important targets of natural selection in human evolution, potentially contributing to the regulatory innovations underlying human-chimpanzee phenotypic differences.

Materials and Methods

Cell culture and differentiation

We utilized two previously characterized human-chimpanzee hybrid iPSC lines (Hy1-30 and Hy1-25) generated through in vitro cell fusion23. Human (two individuals, H20961 and H21976), chimpanzee (two individuals, C3649 and C3647) and human-chimpanzee hybrid induced pluripotent stem cells (iPSCs) were maintained on matrigel in mTeSR1 or mTeSR Plus medium (Stem Cell Technologies cat #85850 or cat #100–0276)23. These lines were confirmed to be free of mycoplasma contamination and their identity was confirmed using the RNA sequencing data generated in this study. The Columbia Stem Cell Core Facility differentiated the iPSCs into multiple cell types including dopaminergic neurons (DA), skeletal myocytes (SKM), and hepatocytes (HEP) using published protocols26,45,95–103. Cranial neural crest cells were differentiated as described24. Primary dental pulp stem cells (DPSC) samples were obtained from two human and two chimpanzee donors as described24.

RNA-sequencing and allele-specific expression analysis

All samples were cryopreserved in liquid nitrogen before RNA extraction104. Cells were gently thawed and then washed with PBS and cell pellets were collected via centrifugation at 1000 RPM for 5 min. RNA extraction was performed using the RNeasy Mini Kit (Qiagen, 74104) and DNase digestion was performed using RNase-Free DNase Set (Qiagen, 79254) following standard protocols. RNA quality was assessed using the Agilent Bioanalyzer RNA Pico assay. Only samples with an RNA integrity number (RIN) greater than or equal to 7 were used to prepare cDNA libraries. Libraries were prepared using TruSeq Stranded mRNA kits (Illumina, 20020594) and sequenced on Illumina HiSeq 4000 to generate paired-end reads. An allele-specific expression pipeline adapted from our previous work was implemented, using dual reference genome alignment to hg38 and pantro6 to eliminate mapping bias23,24. Reads were processed with SeqPrep, mapped using STAR, deduplicated with Picard, and corrected for allelic bias using Hornet. DESeq2 was used for differential expression analysis with FDR correction, defining significant allele-specific expression as FDR < 0.05 with consistent results across both reference genomes23,24,26,105–109. For details, see “Generation of allele-specific count tables” and “Identifying genes with ASE” in the Methods section in Wang et al. (2024)26. For hybrid WGBS samples (DA and CNCC), on average, 48.2%±0.2% of all aligned reads are unassignable using SNPs between human and chimpanzee genome, 25.9%±0.2% of the reads are specific to human, and 25.1%±0.2% of reads are specific to chimpanzee. For hybrid RRBS samples (IPSC, SKM and HEP), on average, 88.4%±0.3% of all aligned reads are unassignable using SNPs between human and chimpanzee genome, 5.3%±0.1% of the reads are specific to human, and 5.3%±0.1% of reads are specific to chimpanzee. The proportion of allele-resolvable reads differs markedly between the two bisulfite assays, and this difference is attributable to fragment length and to the genomic territory each assay interrogates. A read can be assigned to a parental allele only if it spans at least one informative human–chimpanzee SNV. RRBS libraries are generated by MspI digestion and are dominated by short fragments (median aligned length 49–50 bp), and their coverage is concentrated in CpG islands and CpG-dense promoters, which are among the most highly conserved sequences in the genome and therefore carry few interspecies SNVs. Among the 16,845,376 unique CpG positions covered by the unphased hybrid RRBS samples, 65.1% have at least one human–chimpanzee SNV within ±50 bp — the footprint reachable by a typical RRBS read — whereas 90.8% have one within ±150 bp, the footprint of a typical WGBS read pair. Consistent with this, the aligned-length distributions of allele-resolved and unassigned reads are essentially identical in all four WGBS libraries (median 150 bp for both), whereas in all six RRBS libraries allele-resolved reads are substantially longer than unassigned reads (medians 70–75 vs 49–50 bp, respectively).

Bisulfite-seq library preparation, sequencing and mapping

DNA was extracted and processed, and whole-genome bisulfite sequencing (WGBS) and reduced representation bisulfite sequencing (RRBS) was performed at Admera Health Biopharma Services using Illumina Novaseq 6000. WGBS libraries achieved a mean coverage of 9.68× per CpG site, covering 43.6M unique CpG positions in CNCC (hybrid) and 43.4M in CNCC (parental), 41.9M in DA (hybrid, phased) and 38.8M in DA (parental). RRBS libraries achieved a mean coverage of 5.02× per CpG site, covering 5.0M, 4.8M, and 4.5M unique CpG positions in the phased hybrid HEP, IPSC, and SKM libraries respectively, and 3.3M, 3.4M, and 3.2M in the corresponding parental libraries. After phasing the hybrid samples, the total number of CpG sites (union) of all WGBS libraries is 48,994,320 sites, and the union of all CpG sites of RRBS libraries is 7,289,251. The union of all libraries (including WGBS and RRBS) is 49,049,052. We have also included coverage information as a table in Supplemental File 1. To minimize reference mapping bias in cross-species comparisons, reference genomes were first masked at previously identified human-chimpanzee SNV positions using bedtools maskfasta, reducing the potential for allelic mapping preferences26,110. Reference genome preparation was performed using Bismark’s bismark_genome_preparation with Bowtie2 as the aligner, which converts unmethylated cytosines in silico to enable bisulfite-aware alignment111,112. All samples were aligned to both masked human (hg38) and chimpanzee (pantro6) reference genomes to assess mapping quality and species-specific biases. Reads were quality-trimmed and adapter sequences removed using Cutadapt and Trim Galore with default parameters for RRBS libraries, including the removal of artificially filled-in cytosines from the 3’ end of reads113,114. Trimmed reads were aligned using Bismark, and methylation levels were calculated as the ratio of unconverted cytosines to total coverage at each CpG site111. For hybrid samples, SNPsplit was used to assign reads to either the human or chimpanzee allele115. Strand information was collapsed for symmetric CpG sites and genomic annotations were lifted between genomic assemblies hg38 and pantro6 using UCSC LiftOver with reciprocal mapping to ensure orthologous regions116,117.

PCA and hierarchical clustering

Fractional methylation of individual CpG sites was estimated using bismark_methylation_extractor from Bismark111. CpG sites shared across samples were identified using bedtools intersect110. PCA was performed using PCASamples from methylKit across all samples and samples within each cell type118. Euclidean distance matrices were computed across all samples for hierarchical clustering analysis, with visualization performed using the pheatmap package119. Clustering patterns were assessed to validate cell type identity and characterize species-specific methylation signatures.

Beta-binomial regulatory classification framework

Methylation data from human (H) and chimpanzee (C) parental and hybrid samples were analyzed using a beta-binomial generalized linear modeling framework to account for overdispersion inherent to bisulfite sequencing count data. The data included measurements for parental H-genome (ParH), parental C-genome (ParC), hybrid H-allele (HybH), and hybrid C-allele (HybC) for each genomic locus. For each genomic locus (site or annotated region), methylated counts were modeled using two complementary beta-binomial GLM models to separately evaluate cis and trans regulatory effects:

Model I: Cis effect test with methylation data on hybrid alleles logit(π)=β0+β1(Allele) Model II: Trans effect test using System x Allele interaction term with methylation data in both hybrid and parental alleles logit(π)=β0+β1(System)+β2(Allele)+β3(System×Allele) Where π is the probability of methylation, System distinguishes parental from hybrid samples, Allele distinguishes between human and chimpanzee methylation, and the interaction term captures any signals involving the change of allelic differences across generations. Statistical significance was assessed through likelihood ratio tests comparing nested models, testing for system effects, allele effects (cis-regulation), and system-by-allele interactions (trans-regulation). Genomic loci were classified into five categories: conserved, trans-only, cis-only, cis x trans interaction, and cis + trans effects:

Conserved: No significant allelic differences in either generation (neither parental nor hybrid allelic tests significant)

Trans only: Significant allelic difference in parents but not hybrids, indicating trans-acting regulation

Cis only: Significant allelic differences in both systems with no significant interaction, indicating consistent cis-acting effects

Cis × trans: Significant allelic differences in both systems with significant interaction, where effects occur in opposite directions

Cis + trans: Significant allelic differences in both systems with significant interaction, where effects are in the same direction

Regional analyses combined CpG-level methylation across defined genomic intervals (for example, promoters). Regional methylation was quantified as pooled fractional methylation, calculated as the total number of methylated cytosines divided by the total coverage across all CpG sites within each region. Nominal p-values < 0.05 from both models were used to provide a schematic overview of site-level CpG methylation in scatterplots and bar plots, preserving the full dynamic range of the data and providing input features for downstream de novo DMR identification. For reporting significant candidate loci under each regulation category as well as formal inference of region-level regulation, Benjamini–Hochberg false discovery rate (FDR) was calculated to account for multiple testing and to evaluate the robustness of inferred relative contribution of regulatory mechanisms across significance thresholds120. The same classification and visualization steps were employed using FDR-corrected p-values on loci and regions to confirm that the relative contributions of regulatory classifications were not driven by significance cutoffs alone (Figure 2—figure supplement 1 and 2). To assess the robustness of cis-over-trans contributions to methylation with respect to the significance cutoffs, the proportion of significant loci and regions assigned to each regulatory category was estimated across a range of FDR cutoffs (FDR < 0.01, 0.05, 0.1, 0.2, 0.25), confirming that the observed cis dominance over other divergent regulation categories (including trans) was robust to the choice of cutoff (Figure 2—figure supplement 1 and 2).

Changepoint detection for identifying differentially methylated regions with homogeneous cis– or trans-acting methylation divergence

To identify clusters of CpG sites in the same regulation category, we employed the Pruned Exact Linear Time (PELT) algorithm implemented in the changepoint R package to detect shifts in mean methylation scores within each chromosome segment while accounting for a manually specified penalty against over-fitting121–123. We adapted this method to identify regions in the genome where neighboring CpG sites are consistently classified by the previously mentioned beta-binomial model to fall under the same regulatory classification. We first segmented the entire genome into large regions where a break is assigned whenever neighboring CpG sites are more than 5000 base pairs apart. Then, we transform all CpG sites in a segment into binary sequences, where the focal regulation category is 1 and alternative categories are 0. To identify trans-DMRs, we set focal pure trans regulation to 1, and all other regulation categories to 0. To identify cis-DMRs, we set focal pure cis regulation to 1, and all other regulation categories to 0. To identify trans-DMRs specifically for HOMER enrichment inputs, we used a relaxed gradient scoring approach. We set focal pure trans regulation to 1, conserved to 0.75, additive (cis + trans) and interactive (cis x trans) regulation to 0.5 and pure cis regulation category to 0. To determine the optimal penalty parameter, we applied the CROPS (Changepoints for a Range Of Penalties) method, testing penalty values ranging from 0.01 to 2.0. For each penalty range giving the same number of segments, we recorded the number of detected changepoints and corresponding model fit statistics. The optimal penalty for each segment was determined using elbow point detection via the R package kneedle, which identifies the point of maximum distance from line connecting endpoints in the penalty-versus-changepoints curve124. We then applied a homogeneity filter requiring segments to have a minimum target proportion of 0.6, where the target proportion is the fraction of CpG sites within the segment that (i) belong to the target regulatory class under the beta-binomial cis/trans classification and (ii) share a consistent species bias in methylation (all human-biased or all chimpanzee-biased), to ensure biological relevance of identified regions. For segments where fewer than three distinct penalty values yielded different numbers of changepoints, we selected the penalty that maximized the number of segments passing the homogeneity threshold. Segments without detected changepoints were evaluated directly against the homogeneity criterion to assess their potential as single-segment DMRs.

To refine segment boundaries and remove spurious edge effects before merging overlapping DMRs, we implemented a post-processing trimming step that removes segment edges containing sites with regulatory patterns inconsistent with the dominant signal within each segment. This ensures that identified DMRs represent coherent regions of coordinated methylation changes rather than artifacts of the segmentation process. cis-DMRs, trans-DMRs, cis x trans and cis + trans DMRs were identified independently by applying the changepoint detection method separately to each regulatory class.

Identification of cell-type specific DMRs, species-specific DMRs, and cell-type specific species-specific DMRs

Examples of cell-type specific DMRs (HOXA9), species-specific DMRs (CTSF), and cell type-specific species-specific DMRs (LGALS8) were identified in hybrid samples using the DSS-single (Dispersion Shrinkage for Sequencing data)125–128. For cell type-specific DMR identification, we used the formula ∼species + cell type + species:cell type to test for cell type-specific methylation differences, while species-specific DMRs analysis employed ∼species to identify consistent species differences across cell types. Downstream identification of cell-type specific species-specific DMRs identifies the regions where the interaction term species:cell type is significant. The DSS analysis incorporated smoothing across neighboring CpG sites with a smoothing span of 500 bp and required a minimum of 5 CpG sites per region. Statistical significance was determined using Wald tests with FDR < 0.05, and effect size filtering retained regions with methylation difference ≥20%. Integration with gene expression data was performed through Wilcoxon rank-sum tests to identify cell type-specific gene expression patterns associated with cell-type specific DMRs, while permutations (n=1000) were used to assess the significance of coordinated methylation-expression changes in species-specific patterns, which accounts for any background correlation between neighboring genomic features.

Expression-methylation correlation analysis

Human transcript coordinates and gene body annotations were obtained from the hg38.knownGene.gtf.gz file via UCSC Table Browser116, with transcript start positions designated as transcription start sites (TSS). CpG methylation data aggregated within ENCODE-defined promoter regions were analyzed using a beta-binomial classification framework to determine regulatory categories, while corresponding gene expression data underwent parallel hypothesis testing using a negative-binomial model50,129. Genes with the same classification for both methylation and expression—either both cis or both trans—were then analyzed separately. For genes within each regulatory category, the 2 kb surrounding each TSS was split into 10 bins of 200 bp where individual CpG sites were pooled and averaged as fractional methylation values and correlated with expression TPM values. Custom scripts used for binning and correlation are adapted from beta_profile_region.py from CpGtools130.

Sign test selection analysis

To identify coordinated methylation-expression changes potentially under selection, we used a binomial sign test framework that accounts for background probabilities of concordant changes23,24,26,30,131,132. Promoter regions were obtained from the ENCODE Project Consortium129. Canonical transcripts were downloaded from the UCSC Table Browser hg38 assembly. For genes with multiple annotated promoters, we selected the promoter that is the closest to 5’ UTR of the gene, to ensure each gene is assigned one corresponding promoter116. Promoter regions were mapped to canonical transcripts, and functional enrichment using pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) and Human Phenotype Ontology (HPO) was assessed using directional sign tests that leverage the independence of cis-regulatory regions to detect non-random patterns indicative of lineage-specific selection rather than neutral drift1,23,24,26,72–76. For the methylation-only sign test, gene sets with more than 200 genes or fewer than 20 genes with corresponding promoter methylation data were excluded from the analysis. Lineage-specific selection on a gene set was inferred if there was a statistically significant directionality bias in chimpanzee vs. human promoter ASM. For the sign test on genes with putatively repressive methylation—that is, genes where increased methylation within 1 kb of their TSS correspond to decreased expression—we first classified genes as either consistent with repressive methylation (either Ch>Hu methylation paired with Hu>Ch expression, or Hu>Ch methylation paired with Ch>Hu expression) or not consistent. Gene sets with more than 200 genes or fewer than 10 genes with corresponding methylation and expression data were excluded from the analysis. We then reran the sign test using a new background probability defined as the overall fraction of genes with repressive methylation in each cell type. FDRs of the resulting binomial p-values were estimated by randomly permuting the signs of all genes and performing the binomial sign test 104 times. To calculate FDR, we compared each pathway’s empirical binomial p-value against the distribution of permuted p-values. The expected number of false positives was estimated as the average number of pathways reaching that significance level per permutation. FDR was then defined as this expected count of false positives divided by the observed count of pathways with empirical p-values at least as extreme.

cis-DMR sequence variant analysis and trans-DMR motif enrichment analysis

High-confidence human-chimpanzee single nucleotide variants (SNVs) were identified using a dual-reference approach with stringent filtering criteria as described previously26. Allele-specific methylation within different genomic ranges (25bp, 50bp, 100bp) of CpG-altering variants was calculated. Proportions of regions with Hu>Ch and Hu<Ch methylation were calculated for SNVs in three categories: SNVs that disrupt CpG dinucleotides in humans, SNVs that disrupt CpG sites in chimpanzees, and SNVs that lead to no interspecies changes in CpG content between the species. The significance of the difference in relative proportions of Hu>Ch and Hu<Ch of SNVs in the three categories was assessed using a two-proportion test. For proximity tests, the Mann-Whitney U test65 was used to compare the distribution of distances from all cis or all trans- DMRs to the closest CpG SNV. HOMER’s findMotifsGenome.pl was used to test for motif enrichment in the trans-DMR set using the human reference genome hg38 as background 51.